---
title: "Bhandari et al. 2024 | Dexamethasone in LSEC, rebuilt"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

> Written by AlkhemYst-Ai from the run `run_bTqHwtdaRycj`. Each section below is one step of the pipeline, in the order it ran.
>
> Knitting this document runs the analysis again. It expects the run's data files under `data/`, beside the document. The figures it draws should match the ones the run recorded; where a step is named as having produced a figure, that file is in the run's artifacts.

## Setup

```{r step}
# Written by AlchemYst-Crucible, the analysis architecture of AlkhemYst-Ai.
#
# Not written by hand. Every step below corresponds to a component on the
# pipeline canvas, and the settings it runs with are the ones set there.
suppressPackageStartupMessages({
  library(jsonlite)
  library(data.table)
  library(ggplot2)
})

dir.create("artifacts", showWarnings = FALSE, recursive = TRUE)

# An environment, not a list. R copies a list on assignment, so a helper that
# appended to one would write into its own copy and the caller would never see
# it. That is how a run reports success with no metrics at all.
METRICS <- new.env(parent = emptyenv())

metric <- function(node_id, name, value) {
  # Every measurement belongs to the step that made it. The key is what lets
  # the interface show a component's own numbers when it is opened, instead of
  # one pile for the whole run.
  assign(paste0(node_id, "::", name), value, envir = METRICS)
  invisible(value)
}

.wrap_labels <- function(x, width = 45) {
  # str_wrap without depending on stringr being attached.
  vapply(as.character(x), function(s) {
    if (is.na(s) || nchar(s) <= width) return(s)
    paste(strwrap(s, width = width), collapse = "\n")
  }, character(1), USE.NAMES = FALSE)
}

.fit_categories <- function(plot, height) {
  # A FIGURE WITH A ROW PER CATEGORY MUST BE SIZED BY THE CATEGORIES.
  #
  # THE FAULT. An enrichment barplot draws twenty GO terms in whatever frame it
  # is given. "positive regulation of endothelial cell migration" is 56
  # characters, so the label wraps to a second line, the row does not grow to
  # take it, and consecutive terms print through each other. The figure looks
  # finished and cannot be read, which is worse than one that failed, because
  # nothing downstream questions a PNG that exists.
  #
  # This lives here, in the scaffolding, because three runs of telling the
  # coder to size its own figures produced three figures of the same size. A
  # rule the model may skip is not a control; every figure goes through
  # save_fig whether the coder thought about it or not.
  #
  # Anything that throws leaves the plot exactly as it was: a figure drawn at
  # the wrong height is a poor figure, and a figure lost to an error in the
  # code that was trying to improve it is no figure at all.
  tryCatch({
    if (!inherits(plot, "ggplot")) return(list(plot = plot, height = height))
    built <- ggplot2::ggplot_build(plot)
    params <- built$layout$panel_params[[1]]
    labels <- NULL
    if (!is.null(params$y) && !is.null(params$y$get_labels)) {
      labels <- params$y$get_labels()
    }
    labels <- labels[!is.na(labels)]
    # A discrete axis of words, not a numeric scale that happens to print as
    # text: every tick has to be non-numeric before anything is changed.
    if (length(labels) < 3 || !any(is.na(suppressWarnings(as.numeric(labels))))) {
      return(list(plot = plot, height = height))
    }
    longest <- max(nchar(labels))
    if (longest > 24) {
      plot <- plot + ggplot2::scale_y_discrete(labels = .wrap_labels)
      # Wrapping buys legibility at the cost of vertical space, so the rows
      # have to allow for the extra lines it creates.
      lines_each <- ceiling(longest / 45)
      height <- max(height, 0.34 * length(labels) * lines_each + 1.6)
    } else {
      height <- max(height, 0.30 * length(labels) + 1.4)
    }
    list(plot = plot, height = min(height, 24))   # ggsave refuses past ~50in
  }, error = function(e) list(plot = plot, height = height))
}

save_fig <- function(plot, name, width = 7.2, height = 4.8, dpi = 110) {
  # Takes what survminer and its relatives actually return.
  #
  # ggsurvplot() and ggcoxzph() return LISTS, and ggsave() has no method for
  # them: the run dies in grid.draw after the analysis is finished, with an
  # error naming neither the plot nor the package. Printing into a device
  # handles every case and is the only way the risk table and the
  # per-covariate panels survive.
  # NOTHING IS NOT A FIGURE.
  #
  # A step drew with pheatmap and then handed this function a NULL, under a
  # comment claiming save_fig handles pheatmap objects. It does not, and
  # nothing said so: print(NULL) put two bare NULLs in the log, no image was
  # written, and the step reported done. The canvas showed a heatmap step that
  # had run and there was no heatmap. A silent absence is worse than a failure,
  # because a failure gets corrected and this got believed.
  #
  # pheatmap draws to the device and returns its object invisibly, so the
  # figure has to be captured rather than assumed:
  #     p <- pheatmap::pheatmap(mat, silent = TRUE); save_fig(p, "name")
  if (is.null(plot)) {
    stop(sprintf(
      paste0("save_fig('%s') was given NULL, so there is no figure to save. ",
             "Capture the plot object and pass it: for pheatmap, ",
             "p <- pheatmap::pheatmap(mat, silent = TRUE); save_fig(p, '%s')"),
      name, name))
  }
  fitted <- .fit_categories(plot, height)
  plot <- fitted$plot
  height <- fitted$height
  path <- file.path("artifacts", paste0(name, ".png"))
  ok <- tryCatch({
    if (inherits(plot, "ggplot")) {
      ggplot2::ggsave(path, plot = plot, width = width, height = height, dpi = dpi)
    } else {
      grDevices::png(path, width = width * dpi, height = height * dpi, res = dpi)
      print(plot)
      grDevices::dev.off()
    }
    TRUE
  }, error = function(e) {
    message(sprintf("could not save figure '%s': %s", name, conditionMessage(e)))
    FALSE
  })
  invisible(ok)
}

emit_preview <- function(node_id, df, rows = 5) {
  # The table a step produced, as the interface shows it. A transform that
  # reports no metric still has to prove it ran, and its preview is that
  # proof.
  df <- as.data.frame(df)
  n <- min(rows, nrow(df))
  prev <- list(
    rows = nrow(df),
    columns = ncol(df),
    sample_columns = as.list(utils::head(names(df), 500)),
    sample = if (n > 0) unname(lapply(seq_len(n), function(i) {
      as.list(lapply(df[i, , drop = FALSE], function(v) {
        if (is.na(v[1])) NULL else as.character(v[1])
      }))
    })) else list()
  )
  jsonlite::write_json(prev, file.path("artifacts", paste0(node_id, "__preview.json")),
                       auto_unbox = TRUE, null = "null")
  invisible(NULL)
}

annotation_frame <- function(values, cols) {
  # The annotation pheatmap actually accepts.
  #
  # pheatmap matches annotations to the matrix BY NAME, so the thing it wants
  # is a data.frame whose row names are the matrix's column names. Neither
  # obvious wrong answer survives: a named list fails with "incorrect number
  # of dimensions", because a list indexed in two dimensions is not a thing,
  # and a data.frame with no row names fails later and less helpfully inside
  # grid with "'gpar' element 'fill' must not be length 0". Both were
  # reproduced against the real package before this was written.
  #
  # The failure lands after the analysis is finished and the figure is the
  # last thing the step does, so it costs the whole step rather than the
  # picture.
  #
  # THE COLUMN IS NAMED AFTER THE EXPRESSION THAT PRODUCED IT.
  #
  # It used to be named "Group" always. Four annotations built from one sample
  # table and cbind-ed together were then four columns all called Group, and
  # pheatmap died on "Factor levels on variable Group do not match with
  # annotation_colors" -- after edgeR had run, so it cost the heatmap step and
  # every step downstream of it. deparse(substitute()) needs nothing of the
  # caller: annotation_frame(samples$condition, ...) names itself "condition".
  name <- deparse(substitute(values))
  df <- as.data.frame(values, stringsAsFactors = FALSE)
  if (ncol(df) == 1 && is.null(names(values))) {
    name <- sub("^.*\\$", "", name[1])          # samples$condition -> condition
    name <- sub("^.*\\[\\[\"?", "", name)        # meta[["hours"]]    -> hours"]]
    name <- gsub("[^A-Za-z0-9_.]", "", name)
    if (!nzchar(name) || grepl("^[0-9.]", name)) name <- "Group"
    names(df) <- name
  }
  rownames(df) <- cols
  df
}

step_start <- function(node_id, i, total, label) {
  print(sprintf("[[step]] %s start %d/%d ▶ %s", node_id, i, total, label))
  flush.console()
}

step_done <- function(node_id, i, total, label) {
  print(sprintf("[[step]] %s done %d/%d ✓ %s", node_id, i, total, label))
  flush.console()
}

# A step that cannot do its work says so and the run continues. A step that
# stops the program takes every later step with it, including the ones that
# would have worked, and the researcher loses the whole run over one column.
step_skipped <- function(node_id, why) {
  metric(node_id, "skipped", why)
  message(sprintf("%s: %s", node_id, why))
  invisible(NULL)
}
```

## Dataset

```{r n1}
step_start("n1", 1, 7, "Dataset")
DATA_FILE <- "data/lsec_tmt_intensities.csv"
# One reader chosen by extension, because a csv opened as parquet
# fails with a message about magic bytes that names neither.
read_any <- function(path) {
  ext <- tolower(tools::file_ext(path))
  if (ext %in% c("csv", "txt")) data.table::fread(path, data.table = FALSE)
  else if (ext == "tsv") data.table::fread(path, sep = "\t", data.table = FALSE)
  else if (ext == "json") jsonlite::fromJSON(path)
  else if (ext %in% c("parquet", "pq")) arrow::read_parquet(path)
  else data.table::fread(path, data.table = FALSE)
}
df <- read_any(DATA_FILE)
metric("n1", "rows", nrow(df))
metric("n1", "columns", ncol(df))
emit_preview("n1", df)
print(sprintf("loaded %d rows and %d columns from %s", nrow(df), ncol(df), DATA_FILE))
step_done("n1", 1, 7, "Dataset")
```

## edgeR differential expression

Recorded figures from this step: `n4__volcano.png`.

```{r n4}
step_start("n4", 2, 7, "edgeR differential expression")

# ---- load design ------------------------------------------------------------
design_path <- "data/lsec_samples.csv"
if (!file.exists(design_path)) {
  step_skipped("n4", "Design file not found")
  step_done("n4", 2, 7, "edgeR differential expression")
} else {
  samples <- read.csv(design_path, stringsAsFactors = FALSE)

  # ---- sanity checks ---------------------------------------------------------
  req_cols <- c("sample", "condition")
  miss_req <- setdiff(req_cols, colnames(samples))
  if (length(miss_req) > 0) {
    step_skipped("n4", paste("Design missing columns:", paste(miss_req, collapse = ", ")))
    step_done("n4", 2, 7, "edgeR differential expression")
  } else {
    # ---- check gene column in intensity matrix --------------------------------
    if (!"gene" %in% colnames(df)) {
      step_skipped("n4", "'gene' column missing in intensity matrix")
      step_done("n4", 2, 7, "edgeR differential expression")
    } else {
      # ---- match samples -------------------------------------------------------
      sample_cols   <- setdiff(colnames(df), "gene")
      matched_names <- intersect(samples$sample, sample_cols)

      if (length(matched_names) == 0) {
        step_skipped("n4", "No matching sample columns between design and intensity matrix")
        step_done("n4", 2, 7, "edgeR differential expression")
      } else {
        # keep only matched rows and order as in the intensity matrix
        samples_matched <- samples[samples$sample %in% matched_names, ]
        samples_matched <- samples_matched[match(matched_names, samples_matched$sample), ]

        # ---- build count matrix ------------------------------------------------
        counts_mat <- as.matrix(df[, samples_matched$sample, drop = FALSE])
        rownames(counts_mat) <- df$gene

        # collapse duplicated gene identifiers (edgeR requires unique rownames)
        if (any(duplicated(rownames(counts_mat)))) {
          uniq_counts <- rowsum(counts_mat,
                                group = make.unique(rownames(counts_mat)))
          counts_mat <- uniq_counts
        }

        # ---- edgeR pipeline ----------------------------------------------------
        library(edgeR)   ## attached locally for this step only

        y <- DGEList(counts = counts_mat)

        # filter low-count features (min_count from settings)
        min_cnt <- 10
        keep <- filterByExpr(y, min.count = min_cnt)
        if (sum(!keep) == nrow(y)) {
          step_skipped("n4", "All features filtered out by min_count")
          step_done("n4", 2, 7, "edgeR differential expression")
        } else {
          y <- y[keep, , keep.lib.sizes = FALSE]

          # TMM normalization
          y <- calcNormFactors(y, method = "TMM")

                      # design factor
            cond_factor <- factor(samples_matched$condition)
            # estimate dispersion using the condition factor
            design_mat <- model.matrix(~cond_factor)
            y <- estimateDisp(y, design = design_mat)
            ref_level   <- "NT_48h"
            if (!ref_level %in% levels(cond_factor)) {
            step_skipped("n4", paste("Reference level", ref_level, "not present in condition factor"))
            step_done("n4", 2, 7, "edgeR differential expression")
          } else {
            cond_factor <- relevel(cond_factor, ref = ref_level)

            # edgeR exactTest expects a group vector
            y$samples$group <- cond_factor

            # exact test for Dex_48h vs NT_48h
            if (!all(c("Dex_48h", "NT_48h") %in% levels(cond_factor))) {
              step_skipped("n4", "Required groups (Dex_48h, NT_48h) missing")
              step_done("n4", 2, 7, "edgeR differential expression")
            } else {
              et <- exactTest(y, pair = c("NT_48h", "Dex_48h"))

              # full results table
              res_tbl <- topTags(et, n = Inf, adjust.method = "BH", sort.by = "PValue")$table
              res_tbl$gene <- rownames(res_tbl)

              # make a copy for downstream merge step
              n4_res <- res_tbl

              # ---- preview -------------------------------------------------------
              emit_preview("n4", head(res_tbl, 20))

              # ---- metrics --------------------------------------------------------
              alpha <- 0.05
              de_genes <- sum(res_tbl$FDR < alpha, na.rm = TRUE)
              metric("n4_de_genes", "DE genes (FDR < 0.05)", de_genes)

              # ---- volcano plot (evidence) ---------------------------------------
              library(ggplot2)
              volcano <- ggplot(res_tbl,
                                aes(x = logFC, y = -log10(PValue))) +
                geom_point(alpha = 0.4, colour = "steelblue") +
                geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "darkred") +
                geom_hline(yintercept = -log10(alpha), linetype = "dashed", colour = "darkred") +
                labs(title = "edgeR exactTest: Dex_48h vs NT_48h",
                     x = "log2 Fold Change",
                     y = "-log10 P-value") +
                theme_minimal()
              save_fig(volcano, "n4__volcano")

              step_done("n4", 2, 7, "edgeR differential expression")
            }
          }
        }
      }
    }
  }
}
```

## Volcano Plot

Recorded figures from this step: `n5__volcano.png`.

```{r n5}
step_start("n5", 3, 7, "Volcano Plot")

# ---- check that differential expression results exist -------------------------
if (!exists("res_tbl") || nrow(res_tbl) == 0) {
  step_skipped("n5", "Differential expression results (res_tbl) not available")
  step_done("n5", 3, 7, "Volcano Plot")
} else {
  # ---- settings ---------------------------------------------------------------
  lfc_thr   <- 1          # from node settings
  alpha_thr <- 0.05       # from node settings
  label_n   <- 20         # top genes to label

  # ---- prepare data ------------------------------------------------------------
  res_tbl$gene <- as.character(res_tbl$gene)

  # select top genes by smallest FDR (or by absolute logFC if ties)
  top_genes <- head(res_tbl[order(res_tbl$FDR, -abs(res_tbl$logFC)), "gene"], label_n)

  # ---- volcano plot -----------------------------------------------------------
  library(ggplot2)
  library(ggrepel)

  volcano_plot <- ggplot(res_tbl,
                         aes(x = logFC, y = -log10(PValue))) +
    geom_point(alpha = 0.5, colour = "steelblue") +
    geom_vline(xintercept = c(-lfc_thr, lfc_thr), linetype = "dashed", colour = "darkred") +
    geom_hline(yintercept = -log10(alpha_thr), linetype = "dashed", colour = "darkred") +
    geom_text_repel(data = subset(res_tbl, gene %in% top_genes),
                    aes(label = gene),
                    size = 3,
                    box.padding = 0.3,
                    point.padding = 0.2,
                    max.overlaps = Inf) +
    labs(title = "Volcano plot: Dex_48h vs NT_48h",
         x = "log2 Fold Change",
         y = "-log10 P-value") +
    theme_minimal()

  save_fig(volcano_plot, "n5__volcano")

  # ---- optional metric (number of DE genes) ------------------------------------
  de_cnt <- sum(res_tbl$FDR < alpha_thr, na.rm = TRUE)
  metric("n5_de_genes", "DE genes (FDR < 0.05) - volcano step", de_cnt)

  step_done("n5", 3, 7, "Volcano Plot")
}
```

## Expression Heatmap

Recorded figures from this step: `n6__heatmap.png`.

```{r n6}
step_start("n6", 4, 7, "Expression Heatmap")

# ---- load design (samples) ----------------------------------------------------
design_path <- "data/lsec_samples.csv"
if (!file.exists(design_path)) {
  step_skipped("n6", "Design file not found")
  step_done("n6", 4, 7, "Expression Heatmap")
} else {
  samples <- read.csv(design_path, stringsAsFactors = FALSE)

  # sanity check for required columns
  req_cols <- c("sample", "condition")
  miss_req <- setdiff(req_cols, colnames(samples))
  if (length(miss_req) > 0) {
    step_skipped("n6", paste("Design missing columns:", paste(miss_req, collapse = ", ")))
    step_done("n6", 4, 7, "Expression Heatmap")
  } else {
    # ---- match sample columns -------------------------------------------------
    sample_cols   <- setdiff(colnames(df), "gene")
    matched_names <- intersect(samples$sample, sample_cols)

    if (length(matched_names) == 0) {
      step_skipped("n6", "No matching sample columns between design and intensity matrix")
      step_done("n6", 4, 7, "Expression Heatmap")
    } else {
      samples_matched <- samples[samples$sample %in% matched_names, ]
      samples_matched <- samples_matched[match(matched_names, samples_matched$sample), ]

      # ---- build count matrix -------------------------------------------------
      counts_mat <- as.matrix(df[, samples_matched$sample, drop = FALSE])
      rownames(counts_mat) <- df$gene

      # collapse duplicated gene identifiers (edgeR requirement)
      if (any(duplicated(rownames(counts_mat)))) {
        uniq_counts <- rowsum(counts_mat,
                              group = make.unique(rownames(counts_mat)))
        counts_mat <- uniq_counts
      }

      # ---- ensure DE results are present --------------------------------------
      if (!exists("res_tbl") || nrow(res_tbl) == 0) {
        step_skipped("n6", "Differential expression results (res_tbl) not available")
        step_done("n6", 4, 7, "Expression Heatmap")
      } else {
        # ---- select top N DE genes -------------------------------------------
        top_n <- 50   # from node settings
        top_genes <- head(res_tbl[order(res_tbl$FDR, -abs(res_tbl$logFC)), "gene"], top_n)

        # keep only those rows (handle possible missing after collapsing)
        keep_idx <- which(rownames(counts_mat) %in% top_genes)
        if (length(keep_idx) == 0) {
          step_skipped("n6", "No top genes found in expression matrix")
          step_done("n6", 4, 7, "Expression Heatmap")
        } else {
          expr_sub <- counts_mat[keep_idx, , drop = FALSE]

          # ---- transform to log2-CPM ------------------------------------------------
          library(edgeR)   # for cpm()
          cpm_mat <- cpm(expr_sub, log = TRUE, prior.count = 1)

          # optional scaling of rows
          if (TRUE) {   # scale_rows from settings
            row_means <- rowMeans(cpm_mat, na.rm = TRUE)
            row_sds   <- apply(cpm_mat, 1, sd, na.rm = TRUE)
            cpm_mat   <- sweep(cpm_mat, 1, row_means, "-")
            cpm_mat   <- sweep(cpm_mat, 1, row_sds, "/")
          }

          # ---- column annotation -------------------------------------------------
          ann_col <- annotation_frame(samples_matched$condition,
                                      colnames(cpm_mat))

          # ---- draw heatmap ------------------------------------------------------
          library(pheatmap)
          heatmap_obj <- pheatmap(cpm_mat,
                                  annotation_col = ann_col,
                                  show_rownames = nrow(cpm_mat) <= 60, fontsize_row = max(4, min(9, 420 / max(1, nrow(cpm_mat)))),
                                  show_colnames = TRUE,
                                  fontsize_col = 8,
                                  main = "Top DE genes heatmap (log2-CPM)")

          # pheatmap returns a list; the plot is drawn automatically.
          # Save the figure (the object returned by pheatmap is compatible with save_fig)
          save_fig(heatmap_obj, "n6__heatmap")

          # ---- optional preview of the matrix (first few rows) -------------------
          emit_preview("n6", head(cpm_mat, 10))

          step_done("n6", 4, 7, "Expression Heatmap")
        }
      }
    }
  }
}
```

## Enrichment Analysis

Recorded figures from this step: `n9__enrichment_bar.png`, `n9__enrichment_dot.png`.

```{r n9}
step_start("n9", 5, 7, "Enrichment Analysis")

# ---- check for DE results ----------------------------------------------------
if (!exists("res_tbl") || nrow(res_tbl) == 0) {
  step_skipped("n9", "Differential expression results (res_tbl) not available")
  step_done("n9", 5, 7, "Enrichment Analysis")
} else if (!"gene" %in% colnames(res_tbl)) {
  # ---- ensure gene column exists ---------------------------------------------
  step_skipped("n9", "'gene' column missing in differential expression results")
  step_done("n9", 5, 7, "Enrichment Analysis")
} else {
  # ---- settings ---------------------------------------------------------------
  alpha   <- 0.05          # from node settings
  min_gs  <- 10            # minimum gene set size
  ont     <- "BP"          # GO Biological Process
  organism <- "org.Mm.eg.db"   # mouse annotation package

  # ---- prepare gene list -------------------------------------------------------
  sig_genes <- res_tbl$gene[res_tbl$FDR < alpha]
  if (length(sig_genes) == 0) {
    step_skipped("n9", "No significant genes (FDR < 0.05) for enrichment")
    step_done("n9", 5, 7, "Enrichment Analysis")
  } else {
    # ---- load required packages ------------------------------------------------
    suppressPackageStartupMessages({
      library(clusterProfiler)
      library(org.Mm.eg.db)
      library(enrichplot)
      library(ggplot2)
    })

    # ---- run over-representation analysis ---------------------------------------
    ego <- enrichGO(gene          = sig_genes,
                    OrgDb         = org.Mm.eg.db,
                    keyType       = "SYMBOL",
                    ont           = ont,
                    pAdjustMethod = "BH",
                    pvalueCutoff  = alpha,
                    qvalueCutoff  = alpha,
                    minGSSize     = min_gs,
                    readable      = TRUE)

    if (is.null(ego) || nrow(ego) == 0) {
      step_skipped("n9", "Enrichment returned no terms")
      step_done("n9", 5, 7, "Enrichment Analysis")
    } else {
      # ---- preview top enriched terms -----------------------------------------
      emit_preview("n9", head(ego@result, 10))

      # ---- barplot of top terms ------------------------------------------------
      barplot_obj <- barplot(ego, showCategory = 20, title = "GO BP enrichment")
      save_fig(barplot_obj, "n9__enrichment_bar")

      # ---- dotplot as additional evidence --------------------------------------
      dotplot_obj <- dotplot(ego, showCategory = 20, title = "GO BP enrichment")
      save_fig(dotplot_obj, "n9__enrichment_dot")

      # ---- metric: number of enriched terms ------------------------------------
      metric("n9_enriched_terms", "Enriched GO BP terms (FDR < 0.05)", nrow(ego))

      step_done("n9", 5, 7, "Enrichment Analysis")
    }
  }
}
```

## Merge

```{r n7}
step_start("n7", 6, 7, "Merge")

# ---- load published reference ------------------------------------------------
pub_path <- "data/lsec_published_dex48_vs_nt48.csv"
if (!file.exists(pub_path)) {
  step_skipped("n7", "Published reference file not found")
  step_done("n7", 6, 7, "Merge")
} else {
  pub_ref <- read.csv(pub_path, stringsAsFactors = FALSE)

  # ---- sanity checks on reference -------------------------------------------
  req_pub_cols <- c("gene", "published_log2FC", "published_pvalue", "published_fdr")
  miss_pub <- setdiff(req_pub_cols, colnames(pub_ref))
  if (length(miss_pub) > 0) {
    step_skipped("n7", paste("Reference missing columns:", paste(miss_pub, collapse = ", ")))
    step_done("n7", 6, 7, "Merge")
  } else {
    # ---- obtain upstream edgeR results ---------------------------------------
    if (!exists("n4_res")) {
      step_skipped("n7", "Up-stream edgeR results (n4_res) not available")
      step_done("n7", 6, 7, "Merge")
    } else {
      edge_res <- n4_res

      # ---- sanity checks on edgeR results ------------------------------------
      req_edge_cols <- c("gene", "logFC", "PValue", "FDR")
      miss_edge <- setdiff(req_edge_cols, colnames(edge_res))
      if (length(miss_edge) > 0) {
        step_skipped("n7", paste("edgeR results missing columns:", paste(miss_edge, collapse = ", ")))
        step_done("n7", 6, 7, "Merge")
      } else {
        # ---- inner join on gene (1:1 validation) -----------------------------
        merged_tbl <- merge(edge_res, pub_ref,
                            by = "gene",
                            all = FALSE,
                            sort = FALSE)

        # ---- validate 1:1 relationship ---------------------------------------
        dup_gene <- any(duplicated(merged_tbl$gene))
        if (dup_gene) {
          step_skipped("n7", "Duplicate gene identifiers after merge (violates 1:1 validation)")
          step_done("n7", 6, 7, "Merge")
        } else {
          # ---- store for downstream step --------------------------------------
          n7_merged <- merged_tbl   # make available to n8

          # ---- preview ---------------------------------------------------------
          emit_preview("n7", head(merged_tbl, 20))

          # ---- metrics ---------------------------------------------------------
          metric("n7_merged_genes", "Genes successfully merged with reference", nrow(merged_tbl))

          step_done("n7", 6, 7, "Merge")
        }
      }
    }
  }
}
```

## Correlation Test

Recorded figures from this step: `n8__corr_scatter.png`.

```{r n8}
step_start("n8", 7, 7, "Correlation Test")

# ---- ensure merged data is present -------------------------------------------
if (!exists("n7_merged")) {
  step_skipped("n8", "Merged table from n7 not found")
  step_done("n8", 7, 7, "Correlation Test")
} else {
  corr_df <- n7_merged

  # ---- column existence check -------------------------------------------------
  col_a <- "logFC"
  col_b <- "published_log2FC"
  missing_cols <- setdiff(c(col_a, col_b), colnames(corr_df))
  if (length(missing_cols) > 0) {
    step_skipped("n8", paste("Missing columns for correlation:", paste(missing_cols, collapse = ", ")))
    step_done("n8", 7, 7, "Correlation Test")
  } else {
    # ---- compute Pearson correlation -----------------------------------------
    cor_res <- tryCatch(
      cor.test(corr_df[[col_a]], corr_df[[col_b]], method = "pearson"),
      error = function(e) NULL
    )

    if (is.null(cor_res)) {
      step_skipped("n8", "Correlation test failed")
      step_done("n8", 7, 7, "Correlation Test")
    } else {
      # ---- record metrics ------------------------------------------------------
      metric("n8_pearson_r", "Pearson r (logFC vs published_log2FC)", cor_res$estimate)
      metric("n8_pearson_p", "Pearson p-value (logFC vs published_log2FC)", cor_res$p.value)

      # ---- scatter plot with regression line ----------------------------------
      library(ggplot2)
      scatter_plot <- ggplot(corr_df, aes_string(x = col_a, y = col_b)) +
        geom_point(alpha = 0.6, colour = "steelblue") +
        geom_smooth(method = "lm", se = FALSE, colour = "darkred") +
        labs(
          title = "Correlation of edgeR logFC with published logFC",
          subtitle = paste0(
            "r = ", round(cor_res$estimate, 3),
            ", p = ", signif(cor_res$p.value, 3)
          ),
          x = "edgeR log2 Fold Change",
          y = "Published log2 Fold Change"
        ) +
        theme_minimal()

      save_fig(scatter_plot, "n8__corr_scatter")

      step_done("n8", 7, 7, "Correlation Test")
    }
  }
}
```

## Finalize

```{r step-2}
local({
  out <- as.list(METRICS)
  if (length(out) == 0) {
    # Said plainly rather than written as an empty object. A run that measured
    # nothing is nearly always a run where the steps wrote somewhere else, and
    # an empty metrics.json looks identical to a pipeline that had nothing to
    # measure.
    message("no metrics were recorded by any step")
  }
  jsonlite::write_json(out, "artifacts/metrics.json", auto_unbox = TRUE, null = "null")
})
print("pipeline finished")
```

## Metrics recorded by this run

| Metric | Value |
| --- | --- |
| n1::rows | 6028 |
| n8_pearson_r::Pearson r (logFC vs published_log2FC) | 0.9999 |
| n1::columns | 16 |
| n5_de_genes::DE genes (FDR < 0.05) - volcano step | 578 |
| n8_pearson_p::Pearson p-value (logFC vs published_log2FC) | 0 |
| n4_de_genes::DE genes (FDR < 0.05) | 578 |
| n7_merged_genes::Genes successfully merged with reference | 5792 |
| n9_enriched_terms::Enriched GO BP terms (FDR < 0.05) | 1053 |

