---
title: "Microbiome 16S end to end (MiSeq SOP)"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

> Written by AlkhemYst-Ai from the run `run_gVy2RERb1PlM`. Each section below is one step of the pipeline, in the order it ran.
>
> Knitting this document runs the analysis again. It expects the run's data files under `data/`, beside the document. The figures it draws should match the ones the run recorded; where a step is named as having produced a figure, that file is in the run's artifacts.

## Setup

```{r step}
# Written by AlchemYst-Crucible, the analysis architecture of AlkhemYst-Ai.
#
# Not written by hand. Every step below corresponds to a component on the
# pipeline canvas, and the settings it runs with are the ones set there.
suppressPackageStartupMessages({
  library(jsonlite)
  library(data.table)
  library(ggplot2)
})

dir.create("artifacts", showWarnings = FALSE, recursive = TRUE)

# An environment, not a list. R copies a list on assignment, so a helper that
# appended to one would write into its own copy and the caller would never see
# it. That is how a run reports success with no metrics at all.
METRICS <- new.env(parent = emptyenv())

metric <- function(node_id, name, value) {
  # Every measurement belongs to the step that made it. The key is what lets
  # the interface show a component's own numbers when it is opened, instead of
  # one pile for the whole run.
  assign(paste0(node_id, "::", name), value, envir = METRICS)
  invisible(value)
}

.wrap_labels <- function(x, width = 45) {
  # str_wrap without depending on stringr being attached.
  vapply(as.character(x), function(s) {
    if (is.na(s) || nchar(s) <= width) return(s)
    paste(strwrap(s, width = width), collapse = "\n")
  }, character(1), USE.NAMES = FALSE)
}

.fit_categories <- function(plot, height) {
  # A FIGURE WITH A ROW PER CATEGORY MUST BE SIZED BY THE CATEGORIES.
  #
  # THE FAULT. An enrichment barplot draws twenty GO terms in whatever frame it
  # is given. "positive regulation of endothelial cell migration" is 56
  # characters, so the label wraps to a second line, the row does not grow to
  # take it, and consecutive terms print through each other. The figure looks
  # finished and cannot be read, which is worse than one that failed, because
  # nothing downstream questions a PNG that exists.
  #
  # This lives here, in the scaffolding, because three runs of telling the
  # coder to size its own figures produced three figures of the same size. A
  # rule the model may skip is not a control; every figure goes through
  # save_fig whether the coder thought about it or not.
  #
  # Anything that throws leaves the plot exactly as it was: a figure drawn at
  # the wrong height is a poor figure, and a figure lost to an error in the
  # code that was trying to improve it is no figure at all.
  tryCatch({
    if (!inherits(plot, "ggplot")) return(list(plot = plot, height = height))
    built <- ggplot2::ggplot_build(plot)
    params <- built$layout$panel_params[[1]]
    labels <- NULL
    if (!is.null(params$y) && !is.null(params$y$get_labels)) {
      labels <- params$y$get_labels()
    }
    labels <- labels[!is.na(labels)]
    # A discrete axis of words, not a numeric scale that happens to print as
    # text: every tick has to be non-numeric before anything is changed.
    if (length(labels) < 3 || !any(is.na(suppressWarnings(as.numeric(labels))))) {
      return(list(plot = plot, height = height))
    }
    longest <- max(nchar(labels))
    if (longest > 24) {
      plot <- plot + ggplot2::scale_y_discrete(labels = .wrap_labels)
      # Wrapping buys legibility at the cost of vertical space, so the rows
      # have to allow for the extra lines it creates.
      lines_each <- ceiling(longest / 45)
      height <- max(height, 0.34 * length(labels) * lines_each + 1.6)
    } else {
      height <- max(height, 0.30 * length(labels) + 1.4)
    }
    list(plot = plot, height = min(height, 24))   # ggsave refuses past ~50in
  }, error = function(e) list(plot = plot, height = height))
}

save_fig <- function(plot, name, width = 7.2, height = 4.8, dpi = 110) {
  # Takes what survminer and its relatives actually return.
  #
  # ggsurvplot() and ggcoxzph() return LISTS, and ggsave() has no method for
  # them: the run dies in grid.draw after the analysis is finished, with an
  # error naming neither the plot nor the package. Printing into a device
  # handles every case and is the only way the risk table and the
  # per-covariate panels survive.
  # NOTHING IS NOT A FIGURE.
  #
  # A step drew with pheatmap and then handed this function a NULL, under a
  # comment claiming save_fig handles pheatmap objects. It does not, and
  # nothing said so: print(NULL) put two bare NULLs in the log, no image was
  # written, and the step reported done. The canvas showed a heatmap step that
  # had run and there was no heatmap. A silent absence is worse than a failure,
  # because a failure gets corrected and this got believed.
  #
  # pheatmap draws to the device and returns its object invisibly, so the
  # figure has to be captured rather than assumed:
  #     p <- pheatmap::pheatmap(mat, silent = TRUE); save_fig(p, "name")
  if (is.null(plot)) {
    stop(sprintf(
      paste0("save_fig('%s') was given NULL, so there is no figure to save. ",
             "Capture the plot object and pass it: for pheatmap, ",
             "p <- pheatmap::pheatmap(mat, silent = TRUE); save_fig(p, '%s')"),
      name, name))
  }
  fitted <- .fit_categories(plot, height)
  plot <- fitted$plot
  height <- fitted$height
  path <- file.path("artifacts", paste0(name, ".png"))
  ok <- tryCatch({
    if (inherits(plot, "ggplot")) {
      ggplot2::ggsave(path, plot = plot, width = width, height = height, dpi = dpi)
    } else {
      grDevices::png(path, width = width * dpi, height = height * dpi, res = dpi)
      print(plot)
      grDevices::dev.off()
    }
    TRUE
  }, error = function(e) {
    message(sprintf("could not save figure '%s': %s", name, conditionMessage(e)))
    FALSE
  })
  invisible(ok)
}

emit_preview <- function(node_id, df, rows = 5) {
  # The table a step produced, as the interface shows it. A transform that
  # reports no metric still has to prove it ran, and its preview is that
  # proof.
  df <- as.data.frame(df)
  n <- min(rows, nrow(df))
  prev <- list(
    rows = nrow(df),
    columns = ncol(df),
    sample_columns = as.list(utils::head(names(df), 500)),
    sample = if (n > 0) unname(lapply(seq_len(n), function(i) {
      as.list(lapply(df[i, , drop = FALSE], function(v) {
        if (is.na(v[1])) NULL else as.character(v[1])
      }))
    })) else list()
  )
  jsonlite::write_json(prev, file.path("artifacts", paste0(node_id, "__preview.json")),
                       auto_unbox = TRUE, null = "null")
  invisible(NULL)
}

annotation_frame <- function(values, cols) {
  # The annotation pheatmap actually accepts.
  #
  # pheatmap matches annotations to the matrix BY NAME, so the thing it wants
  # is a data.frame whose row names are the matrix's column names. Neither
  # obvious wrong answer survives: a named list fails with "incorrect number
  # of dimensions", because a list indexed in two dimensions is not a thing,
  # and a data.frame with no row names fails later and less helpfully inside
  # grid with "'gpar' element 'fill' must not be length 0". Both were
  # reproduced against the real package before this was written.
  #
  # The failure lands after the analysis is finished and the figure is the
  # last thing the step does, so it costs the whole step rather than the
  # picture.
  #
  # THE COLUMN IS NAMED AFTER THE EXPRESSION THAT PRODUCED IT.
  #
  # It used to be named "Group" always. Four annotations built from one sample
  # table and cbind-ed together were then four columns all called Group, and
  # pheatmap died on "Factor levels on variable Group do not match with
  # annotation_colors" -- after edgeR had run, so it cost the heatmap step and
  # every step downstream of it. deparse(substitute()) needs nothing of the
  # caller: annotation_frame(samples$condition, ...) names itself "condition".
  name <- deparse(substitute(values))
  df <- as.data.frame(values, stringsAsFactors = FALSE)
  if (ncol(df) == 1 && is.null(names(values))) {
    name <- sub("^.*\\$", "", name[1])          # samples$condition -> condition
    name <- sub("^.*\\[\\[\"?", "", name)        # meta[["hours"]]    -> hours"]]
    name <- gsub("[^A-Za-z0-9_.]", "", name)
    if (!nzchar(name) || grepl("^[0-9.]", name)) name <- "Group"
    names(df) <- name
  }
  rownames(df) <- cols
  df
}

# The command-line tools a step may run, and the only way to run them.
#
# A generated program may not call the shell (system, system2 and the rest
# are refused by the harness), because a program that can run anything can
# read, write or send anything. Sequencing work still needs a handful of
# standard tools that have no R equivalent: cutadapt to remove primers,
# FastQC for the read report, MAFFT and FastTree for the phylogeny, PICRUSt2
# for predicted function. So there is one door, here in the scaffold the
# platform owns: a fixed list of installed programs, run by absolute path with
# their arguments passed one by one and never through a shell, inside the
# run's own sandbox (no network), with a time limit. What each tool printed is
# kept as an artifact of the step that ran it.
.RUN_TOOLS <- list(
  cutadapt = "/usr/bin/cutadapt",
  fastqc = "/usr/bin/fastqc",
  mafft = "/usr/bin/mafft",
  FastTree = "/usr/bin/fasttree",
  picrust2_pipeline.py = "/opt/picrust2/bin/picrust2_pipeline.py",
  pathway_pipeline.py = "/opt/picrust2/bin/pathway_pipeline.py",
  add_descriptions.py = "/opt/picrust2/bin/add_descriptions.py"
)

run_tool <- function(tool, args = character(), node_id = NULL, stdout_file = NULL, timeout = 7200) {
  path <- .RUN_TOOLS[[tool]]
  if (is.null(path)) {
    stop(sprintf("run_tool: '%s' is not one of the tools a step may run (%s)", tool,
                 paste(names(.RUN_TOOLS), collapse = ", ")), call. = FALSE)
  }
  if (!file.exists(path)) stop(sprintf("run_tool: %s is not installed on this runner", tool), call. = FALSE)
  args <- as.character(unlist(args))
  # PICRUSt2 calls its own helpers (EPA-ng, gappa, HMMER, its R) by name
  env <- if (startsWith(path, "/opt/picrust2/")) c("current", PATH = paste("/opt/picrust2/bin", Sys.getenv("PATH"), sep = ":")) else "current"
  res <- processx::run(path, args, error_on_status = FALSE, timeout = timeout, env = env,
                       stdout = if (is.null(stdout_file)) "|" else stdout_file, stderr = "|")
  if (!is.null(node_id)) {
    log <- c(sprintf("$ %s %s", tool, paste(shQuote(args), collapse = " ")), "",
             if (is.null(stdout_file)) res$stdout else sprintf("(output written to %s)", stdout_file), res$stderr)
    writeLines(log, file.path("artifacts", sprintf("%s__%s.log", node_id, gsub("[^A-Za-z0-9]+", "_", tool))))
  }
  if (isTRUE(res$timeout)) stop(sprintf("run_tool: %s did not finish within %d seconds", tool, timeout), call. = FALSE)
  if (!identical(res$status, 0L)) {
    tail_err <- paste(utils::tail(strsplit(res$stderr, "\n")[[1]], 12), collapse = "\n")
    stop(sprintf("run_tool: %s exited with status %s\n%s", tool, res$status, tail_err), call. = FALSE)
  }
  invisible(res)
}

step_start <- function(node_id, i, total, label) {
  print(sprintf("[[step]] %s start %d/%d ▶ %s", node_id, i, total, label))
  flush.console()
}

step_done <- function(node_id, i, total, label) {
  print(sprintf("[[step]] %s done %d/%d ✓ %s", node_id, i, total, label))
  flush.console()
}

# A step that cannot do its work says so and the run continues. A step that
# stops the program takes every later step with it, including the ones that
# would have worked, and the researcher loses the whole run over one column.
step_skipped <- function(node_id, why) {
  metric(node_id, "skipped", why)
  message(sprintf("%s: %s", node_id, why))
  invisible(NULL)
}
```

## Dataset

```{r n0}
step_start("n0", 1, 15, "Dataset")
DATA_ARCHIVE <- "data/miseq_sop_16S.zip"
TABLE_FILES <- list.files("data", pattern = "[.](csv|tsv|txt)([.]gz)?$", full.names = TRUE)
READ_FILES <- sort(list.files("data", pattern = "[.](fastq|fq)([.]gz)?$", full.names = TRUE))
read_any <- function(path) {
  ext <- tolower(tools::file_ext(path))
  if (ext == "tsv") data.table::fread(path, sep = "\t", data.table = FALSE)
  else data.table::fread(path, data.table = FALSE)
}
DATA_FILE <- if (length(TABLE_FILES)) TABLE_FILES[which.max(file.size(TABLE_FILES))] else DATA_ARCHIVE
df <- if (length(TABLE_FILES)) read_any(DATA_FILE) else data.frame()
metric("n0", "rows", nrow(df))
metric("n0", "columns", ncol(df))
metric("n0", "read_files", length(READ_FILES))
if (nrow(df)) emit_preview("n0", df)
print(sprintf("archive %s: %d read files, sample table %s (%d rows)", DATA_ARCHIVE,
              length(READ_FILES), basename(DATA_FILE), nrow(df)))
step_done("n0", 1, 15, "Dataset")
```

## Read Quality Report

Recorded figures from this step: `n1__quality_forward.png`, `n1__quality_reverse.png`, `n1__reads_per_sample.png`.

```{r n1}
step_start("n1", 1, 4, "Read Quality Report")
settings <- list(`r1_pattern` = "_R1", `r2_pattern` = "_R2", `n_reads` = 5000, `run_fastqc` = TRUE)   # this step's settings from the canvas (tested program, pinned by the platform)
# Read Quality Report: the reads as they arrive, per sample and per position.
if (!exists("mb_reads")) {
  files <- if (exists("READ_FILES") && length(READ_FILES)) READ_FILES else
    sort(list.files("data", pattern = "[.](fastq|fq)([.]gz)?$", full.names = TRUE))
  r1 <- if (length(settings$r1_pattern) && nzchar(settings$r1_pattern)) settings$r1_pattern else "_R1"
  r2 <- if (length(settings$r2_pattern) && nzchar(settings$r2_pattern)) settings$r2_pattern else "_R2"
  fwd <- files[grepl(r1, basename(files), fixed = TRUE)]
  rev <- files[grepl(r2, basename(files), fixed = TRUE)]
  if (!length(fwd)) { fwd <- files; rev <- character() }
  ids <- if (exists("df") && nrow(df)) as.character(df[[1]]) else character()
  sample_of <- function(b) {
    hit <- ids[startsWith(b, paste0(ids, "_")) | startsWith(b, paste0(ids, ".")) | startsWith(b, paste0(ids, "-"))]
    if (length(hit)) hit[which.max(nchar(hit))] else sub("[_.].*$", "", b)
  }
  mb_reads <- data.frame(sample = vapply(basename(fwd), sample_of, character(1), USE.NAMES = FALSE),
                         fwd = fwd, stringsAsFactors = FALSE)
  mb_reads$rev <- if (length(rev)) rev[match(sub(r1, r2, basename(fwd), fixed = TRUE), basename(rev))] else NA_character_
}
if (!nrow(mb_reads)) {
  step_skipped("n1", "no FASTQ files were found in the dataset")
} else {
  paired <- all(!is.na(mb_reads$rev))
  count_reads <- function(f) length(ShortRead::readFastq(f))
  mb_reads$reads <- vapply(mb_reads$fwd, count_reads, numeric(1))
  metric("n1", "samples", nrow(mb_reads))
  metric("n1", "paired_end", paired)
  metric("n1", "median_reads_per_sample", stats::median(mb_reads$reads))
  metric("n1", "min_reads_per_sample", min(mb_reads$reads))
  emit_preview("n1", data.frame(sample = mb_reads$sample, reads = mb_reads$reads,
    forward = basename(mb_reads$fwd), reverse = basename(mb_reads$rev)))
  # per-position quality from a sample of reads, the way DADA2 draws it
  n <- if (length(settings$n_reads)) settings$n_reads else 5000
  save_fig(dada2::plotQualityProfile(mb_reads$fwd, aggregate = TRUE, n = n) +
    ggplot2::ggtitle("Forward reads, all samples"), "n1__quality_forward")
  if (paired) save_fig(dada2::plotQualityProfile(mb_reads$rev, aggregate = TRUE, n = n) +
    ggplot2::ggtitle("Reverse reads, all samples"), "n1__quality_reverse")
  save_fig(ggplot2::ggplot(mb_reads, ggplot2::aes(stats::reorder(sample, reads), reads)) +
    ggplot2::geom_col(fill = "#C27840") + ggplot2::coord_flip() +
    ggplot2::labs(x = NULL, y = "reads"), "n1__reads_per_sample")
  # FastQC, the report every sequencing core sends: one HTML page per file
  if (!isFALSE(settings$run_fastqc)) {
    out <- tempfile("fastqc_"); dir.create(out)
    run_tool("fastqc", c("--outdir", out, "--quiet", "--threads", "2", mb_reads$fwd, stats::na.omit(mb_reads$rev)), node_id = "n1")
    status <- list()
    for (z in list.files(out, pattern = "_fastqc[.]zip$", full.names = TRUE)) {
      inner <- utils::unzip(z, list = TRUE)$Name
      s <- utils::read.delim(unz(z, grep("summary.txt$", inner, value = TRUE)[1]), header = FALSE, col.names = c("status", "module", "file"))
      status[[length(status) + 1]] <- s
    }
    for (h in list.files(out, pattern = "_fastqc[.]html$", full.names = TRUE)) {
      writeLines(readLines(h, warn = FALSE), file.path("artifacts", paste0("n1__", sub("[.]html$", "", basename(h)), ".html")))
    }
    if (length(status)) {
      st <- do.call(rbind, status)
      tab <- as.data.frame.matrix(table(st$module, st$status))
      tab <- data.frame(module = rownames(tab), tab, check.names = FALSE)
      metric("n1", "fastqc_modules_failed", sum(st$status == "FAIL"))
      metric("n1", "fastqc_modules_warned", sum(st$status == "WARN"))
      data.table::fwrite(tab, "artifacts/n1__fastqc_summary.csv")
    }
  }
}
step_done("n1", 1, 4, "Read Quality Report")
```

## Primer Trimming (Cutadapt)

Recorded figures from this step: `n2__reads_before_after.png`.

```{r n2}
step_start("n2", 2, 4, "Primer Trimming (Cutadapt)")
settings <- list(`primer_forward` = "GTGCCAGCMGCCGCGGTAA", `primer_reverse` = "GGACTACHVGGGTWTCTAAT", `error_rate` = 0.1, `min_length` = 50, `discard_untrimmed` = TRUE)   # this step's settings from the canvas (tested program, pinned by the platform)
# Primer Trimming (Cutadapt): remove the amplification primers from both ends,
# and their reverse complements where a short amplicon reads through.
if (!exists("mb_reads")) {
  files <- if (exists("READ_FILES") && length(READ_FILES)) READ_FILES else
    sort(list.files("data", pattern = "[.](fastq|fq)([.]gz)?$", full.names = TRUE))
  fwd <- files[grepl("_R1", basename(files), fixed = TRUE)]
  rev <- files[grepl("_R2", basename(files), fixed = TRUE)]
  if (!length(fwd)) { fwd <- files; rev <- character() }
  ids <- if (exists("df") && nrow(df)) as.character(df[[1]]) else character()
  sample_of <- function(b) {
    hit <- ids[startsWith(b, paste0(ids, "_")) | startsWith(b, paste0(ids, ".")) | startsWith(b, paste0(ids, "-"))]
    if (length(hit)) hit[which.max(nchar(hit))] else sub("[_.].*$", "", b)
  }
  mb_reads <- data.frame(sample = vapply(basename(fwd), sample_of, character(1), USE.NAMES = FALSE),
                         fwd = fwd, stringsAsFactors = FALSE)
  mb_reads$rev <- if (length(rev)) rev[match(sub("_R1", "_R2", basename(fwd), fixed = TRUE), basename(rev))] else NA_character_
}
fwd_primer <- toupper(gsub("\\s", "", settings$primer_forward))
rev_primer <- toupper(gsub("\\s", "", settings$primer_reverse))
if (!nzchar(fwd_primer)) {
  step_skipped("n2", "no forward primer was given")
} else {
  rc <- function(p) as.character(Biostrings::reverseComplement(Biostrings::DNAString(p)))
  paired <- all(!is.na(mb_reads$rev)) && nzchar(rev_primer)
  out <- tempfile("cutadapt_"); dir.create(out)
  rows <- list()
  trimmed <- mb_reads
  for (i in seq_len(nrow(mb_reads))) {
    f1 <- file.path(out, basename(mb_reads$fwd[i]))
    rep <- file.path(out, paste0(mb_reads$sample[i], ".json"))
    args <- c("-g", fwd_primer, "-e", as.character(settings$error_rate), "--minimum-length", as.character(settings$min_length),
              "-j", "2", "--json", rep, "-o", f1)
    if (paired) {
      f2 <- file.path(out, basename(mb_reads$rev[i]))
      args <- c(args, "-a", rc(rev_primer), "-G", rev_primer, "-A", rc(fwd_primer), "-p", f2)
    } else if (nzchar(rev_primer)) {
      args <- c(args, "-a", rc(rev_primer))
    }
    if (!isFALSE(settings$discard_untrimmed)) args <- c(args, "--discard-untrimmed")
    args <- c(args, "-n", "2", mb_reads$fwd[i], if (paired) mb_reads$rev[i])
    run_tool("cutadapt", args, node_id = if (i == 1) "n2" else NULL)
    j <- jsonlite::fromJSON(rep)
    rc_in <- j$read_counts$input
    rc_out <- j$read_counts$output
    with_primer <- j$read_counts$read1_with_adapter
    rows[[i]] <- data.frame(sample = mb_reads$sample[i], reads_in = rc_in, reads_out = rc_out,
                            with_primer = if (is.null(with_primer)) NA else with_primer)
    trimmed$fwd[i] <- f1
    if (paired) trimmed$rev[i] <- f2
  }
  tab <- do.call(rbind, rows)
  found <- sum(tab$with_primer, na.rm = TRUE) / max(sum(tab$reads_in), 1)
  # Reads that were trimmed before upload, or sequenced from past the primer,
  # carry no primer. Discarding every read without one would then discard the
  # dataset, so the step says so and the original reads go on, and what it
  # reports is what goes on, not what cutadapt would have kept.
  if (found < 0.2) {
    tab$reads_out <- tab$reads_in
    tab$note <- "no primer in the reads: passed on unchanged"
  }
  tab$kept_pct <- round(100 * tab$reads_out / pmax(tab$reads_in, 1), 1)
  metric("n2", "reads_in", sum(tab$reads_in))
  metric("n2", "reads_out", sum(tab$reads_out))
  metric("n2", "primer_found_pct", round(100 * found, 1))
  emit_preview("n2", tab)
  if (found < 0.2) {
    metric("n2", "primers_absent", TRUE)
    message("n2: the primer was found in ", round(100 * found, 1), "% of reads; the reads look already trimmed, so the original reads are used")
  } else {
    mb_raw_reads <- mb_reads
    mb_reads <- trimmed
    mb_reads <- mb_reads[file.exists(mb_reads$fwd) & tab$reads_out > 0, ]
  }
  long <- rbind(data.frame(sample = tab$sample, stage = "before", reads = tab$reads_in),
                data.frame(sample = tab$sample, stage = "after", reads = tab$reads_out))
  save_fig(ggplot2::ggplot(long, ggplot2::aes(sample, reads, fill = stage)) +
    ggplot2::geom_col(position = "dodge") + ggplot2::coord_flip() +
    ggplot2::scale_fill_manual(values = c(before = "#9A8F84", after = "#C27840")) +
    ggplot2::labs(x = NULL, y = "reads", fill = NULL), "n2__reads_before_after")
}
step_done("n2", 2, 4, "Primer Trimming (Cutadapt)")
```

## Filter & Trim Reads

Recorded figures from this step: `n3__truncation_choice.png`.

```{r n3}
step_start("n3", 3, 4, "Filter & Trim Reads")
settings <- list(`amplicon` = "16S V4", `amplicon_length` = 253, `auto_truncation` = TRUE, `quality_threshold` = 25, `min_overlap` = 20, `trunc_len_forward` = 240, `trunc_len_reverse` = 160, `max_ee_forward` = 2, `max_ee_reverse` = 2, `min_length` = 50)   # this step's settings from the canvas (tested program, pinned by the platform)
# Filter & Trim Reads: quality filtering, with the truncation lengths chosen
# from the reads' own quality profile unless they are set by hand.
if (!exists("mb_reads")) {
  files <- if (exists("READ_FILES") && length(READ_FILES)) READ_FILES else
    sort(list.files("data", pattern = "[.](fastq|fq)([.]gz)?$", full.names = TRUE))
  fwd <- files[grepl("_R1", basename(files), fixed = TRUE)]
  rev <- files[grepl("_R2", basename(files), fixed = TRUE)]
  if (!length(fwd)) { fwd <- files; rev <- character() }
  ids <- if (exists("df") && nrow(df)) as.character(df[[1]]) else character()
  sample_of <- function(b) {
    hit <- ids[startsWith(b, paste0(ids, "_")) | startsWith(b, paste0(ids, ".")) | startsWith(b, paste0(ids, "-"))]
    if (length(hit)) hit[which.max(nchar(hit))] else sub("[_.].*$", "", b)
  }
  mb_reads <- data.frame(sample = vapply(basename(fwd), sample_of, character(1), USE.NAMES = FALSE),
                         fwd = fwd, stringsAsFactors = FALSE)
  mb_reads$rev <- if (length(rev)) rev[match(sub("_R1", "_R2", basename(fwd), fixed = TRUE), basename(rev))] else NA_character_
}
paired <- all(!is.na(mb_reads$rev))
region <- settings$amplicon
amplicon_len <- switch(region, "16S V4" = 253, "16S V3-V4" = 465, "16S V4-V5" = 411, "18S V4" = 420, "18S V9" = 130,
                       "ITS" = NA, as.numeric(settings$amplicon_length))
# The quality at each position, lower quartile, from a sample of reads
q_profile <- function(files, n = 20000) {
  each <- max(200, ceiling(n / length(files)))
  qs <- lapply(files, function(f) {
    s <- ShortRead::FastqSampler(f, each); fq <- ShortRead::yield(s); close(s)
    methods::as(Biostrings::quality(fq), "matrix")
  })
  width <- max(vapply(qs, ncol, numeric(1)))
  qs <- lapply(qs, function(m) { if (ncol(m) < width) cbind(m, matrix(NA, nrow(m), width - ncol(m))) else m })
  m <- do.call(rbind, qs)
  data.frame(position = seq_len(ncol(m)), q25 = apply(m, 2, stats::quantile, 0.25, na.rm = TRUE),
             median = apply(m, 2, stats::median, na.rm = TRUE))
}
choose_len <- function(p, threshold) {
  bad <- which(p$q25 < threshold)
  if (!length(bad)) nrow(p) else max(bad[1] - 1, 50)
}
pf <- q_profile(mb_reads$fwd)
pr <- if (paired) q_profile(mb_reads$rev) else NULL
threshold <- as.numeric(settings$quality_threshold)
if (identical(region, "ITS")) {
  # ITS lengths vary between taxa: truncating to one length would remove whole taxa
  tl <- c(0, 0); how <- "none (ITS lengths vary)"
} else if (isTRUE(settings$auto_truncation)) {
  tl <- c(choose_len(pf, threshold), if (paired) choose_len(pr, threshold) else 0)
  how <- sprintf("from the quality profile (lower quartile above Q%s)", threshold)
  # paired reads must still overlap across the amplicon to be merged
  need <- if (paired && !is.na(amplicon_len)) amplicon_len + as.numeric(settings$min_overlap) else 0
  if (paired && need > 0 && sum(tl) < need) {
    short <- need - sum(tl)
    grow_f <- min(nrow(pf) - tl[1], ceiling(short / 2)); tl[1] <- tl[1] + grow_f
    tl[2] <- min(nrow(pr), tl[2] + (need - sum(tl)))
    how <- paste0(how, ", lengthened to keep a ", settings$min_overlap, "-base overlap over the ", amplicon_len, "-base amplicon")
    if (sum(tl) < need) message("n3: the reads are too short to overlap across this amplicon; merging will lose most pairs")
  }
} else {
  tl <- c(as.numeric(settings$trunc_len_forward), if (paired) as.numeric(settings$trunc_len_reverse) else 0)
  how <- "set by hand"
}
metric("n3", "trunc_len_forward", tl[1])
if (paired) metric("n3", "trunc_len_reverse", tl[2])
metric("n3", "truncation", how)
filt_dir <- tempfile("filtered_"); dir.create(filt_dir)
filt_f <- file.path(filt_dir, paste0(mb_reads$sample, "_F_filt.fastq.gz"))
filt_r <- if (paired) file.path(filt_dir, paste0(mb_reads$sample, "_R_filt.fastq.gz")) else NULL
max_ee <- c(as.numeric(settings$max_ee_forward), as.numeric(settings$max_ee_reverse))
out <- if (paired) {
  dada2::filterAndTrim(mb_reads$fwd, filt_f, mb_reads$rev, filt_r, truncLen = tl, maxN = 0, maxEE = max_ee,
                       truncQ = 2, minLen = as.numeric(settings$min_length), rm.phix = TRUE, compress = TRUE, multithread = 2)
} else {
  dada2::filterAndTrim(mb_reads$fwd, filt_f, truncLen = tl[1], maxN = 0, maxEE = max_ee[1],
                       truncQ = 2, minLen = as.numeric(settings$min_length), rm.phix = TRUE, compress = TRUE, multithread = 2)
}
mb_filter_counts <- data.frame(sample = mb_reads$sample, input = out[, 1], filtered = out[, 2])
keep <- file.exists(filt_f) & out[, 2] > 0
mb_filt <- data.frame(sample = mb_reads$sample[keep], fwd = filt_f[keep],
                      rev = if (paired) filt_r[keep] else NA_character_, stringsAsFactors = FALSE)
metric("n3", "reads_in", sum(out[, 1]))
metric("n3", "reads_kept", sum(out[, 2]))
metric("n3", "reads_kept_pct", round(100 * sum(out[, 2]) / max(sum(out[, 1]), 1), 1))
if (any(!keep)) metric("n3", "samples_emptied", paste(mb_reads$sample[!keep], collapse = ", "))
emit_preview("n3", transform(mb_filter_counts, kept_pct = round(100 * filtered / pmax(input, 1), 1)))
prof <- rbind(transform(pf, read = "forward"), if (paired) transform(pr, read = "reverse"))
cut <- data.frame(read = c("forward", if (paired) "reverse"), at = tl[seq_len(if (paired) 2 else 1)])
save_fig(ggplot2::ggplot(prof, ggplot2::aes(position, q25)) + ggplot2::geom_line(colour = "#C27840") +
  ggplot2::geom_line(ggplot2::aes(y = median), colour = "#201C18", linetype = "dashed") +
  ggplot2::geom_hline(yintercept = threshold, colour = "#9A8F84") +
  { if (all(cut$at > 0)) ggplot2::geom_vline(data = cut, ggplot2::aes(xintercept = at), colour = "#A13D2D") } +
  ggplot2::facet_wrap(~read, ncol = 1) + ggplot2::labs(y = "quality (lower quartile, and median dashed)"),
  "n3__truncation_choice")
step_done("n3", 3, 4, "Filter & Trim Reads")
```

## ASV Inference (DADA2)

Recorded figures from this step: `n4__asv_lengths.png`, `n4__error_rates.png`, `n4__reads_through_the_steps.png`.

```{r n4}
step_start("n4", 4, 4, "ASV Inference (DADA2)")
settings <- list(`sample_col` = "sample", `pool` = "false", `min_overlap` = 12, `chimera_method` = "consensus", `learn_bases` = 100000000, `seed` = 100)   # this step's settings from the canvas (tested program, pinned by the platform)
# ASV Inference (DADA2): learn the run's error rates, denoise each sample into
# exact sequence variants, merge the pairs, and remove chimeras.
if (!exists("mb_filt")) {
  step_skipped("n4", "run Filter & Trim Reads first: DADA2 needs filtered reads")
} else {
  set.seed(as.integer(settings$seed))
  paired <- all(!is.na(mb_filt$rev))
  pool <- switch(settings$pool, "pseudo" = "pseudo", "true" = TRUE, FALSE)
  nbases <- as.numeric(settings$learn_bases)
  err_f <- dada2::learnErrors(mb_filt$fwd, nbases = nbases, multithread = 2, verbose = FALSE)
  dd_f <- dada2::dada(mb_filt$fwd, err = err_f, pool = pool, multithread = 2, verbose = FALSE)
  if (inherits(dd_f, "dada")) dd_f <- list(dd_f)
  names(dd_f) <- mb_filt$sample
  if (paired) {
    err_r <- dada2::learnErrors(mb_filt$rev, nbases = nbases, multithread = 2, verbose = FALSE)
    dd_r <- dada2::dada(mb_filt$rev, err = err_r, pool = pool, multithread = 2, verbose = FALSE)
    if (inherits(dd_r, "dada")) dd_r <- list(dd_r)
    names(dd_r) <- mb_filt$sample
    merged <- dada2::mergePairs(dd_f, mb_filt$fwd, dd_r, mb_filt$rev, minOverlap = as.integer(settings$min_overlap), verbose = FALSE)
    if (is.data.frame(merged)) merged <- list(merged)
    names(merged) <- mb_filt$sample
    seqtab_all <- dada2::makeSequenceTable(merged)
  } else {
    seqtab_all <- dada2::makeSequenceTable(dd_f)
  }
  seqtab <- dada2::removeBimeraDenovo(seqtab_all, method = settings$chimera_method, multithread = 2, verbose = FALSE)
  rownames(seqtab) <- mb_filt$sample
  getN <- function(x) sum(dada2::getUniques(x))
  mb_track <- data.frame(sample = mb_filt$sample,
    denoised_forward = vapply(dd_f, getN, numeric(1)),
    merged = if (paired) vapply(merged, getN, numeric(1)) else vapply(dd_f, getN, numeric(1)),
    non_chimeric = rowSums(seqtab))
  if (exists("mb_filter_counts")) mb_track <- merge(mb_filter_counts, mb_track, by = "sample", all.y = TRUE)
  # ASVs are named by abundance; the sequence stays with each name
  ord <- order(-colSums(seqtab))
  seqtab <- seqtab[, ord, drop = FALSE]
  asv_seqs <- Biostrings::DNAStringSet(colnames(seqtab))
  names(asv_seqs) <- paste0("ASV", seq_along(asv_seqs))
  colnames(seqtab) <- names(asv_seqs)
  meta <- if (exists("df") && nrow(df)) df else data.frame(sample = rownames(seqtab))
  scol <- settings$sample_col
  if (!length(scol) || !nzchar(scol) || !(scol %in% names(meta))) {
    scol <- names(meta)[which.max(vapply(meta, function(v) sum(as.character(v) %in% rownames(seqtab)), numeric(1)))]
  }
  meta <- meta[!duplicated(meta[[scol]]), , drop = FALSE]
  rownames(meta) <- as.character(meta[[scol]])
  meta <- meta[rownames(seqtab), , drop = FALSE]
  rownames(meta) <- rownames(seqtab)
  meta[[scol]] <- rownames(seqtab)
  ps <- phyloseq::phyloseq(phyloseq::otu_table(seqtab, taxa_are_rows = FALSE),
                           phyloseq::sample_data(meta), phyloseq::refseq(asv_seqs))
  mb_sample_col <- scol
  metric("n4", "asvs", ncol(seqtab))
  metric("n4", "samples", nrow(seqtab))
  metric("n4", "reads_non_chimeric", sum(seqtab))
  metric("n4", "chimeric_reads_pct", round(100 * (1 - sum(seqtab) / max(sum(seqtab_all), 1)), 2))
  if ("input" %in% names(mb_track)) metric("n4", "reads_retained_pct", round(100 * sum(mb_track$non_chimeric) / max(sum(mb_track$input), 1), 1))
  emit_preview("n4", mb_track)
  save_fig(dada2::plotErrors(err_f, nominalQ = TRUE) + ggplot2::ggtitle("Error rates, forward reads"), "n4__error_rates", height = 6)
  lens <- data.frame(length = Biostrings::width(asv_seqs), reads = colSums(seqtab))
  save_fig(ggplot2::ggplot(lens, ggplot2::aes(length, weight = reads)) + ggplot2::geom_histogram(binwidth = 1, fill = "#C27840") +
    ggplot2::labs(x = "ASV length (bases)", y = "reads"), "n4__asv_lengths")
  long <- stats::reshape(mb_track, direction = "long", varying = setdiff(names(mb_track), "sample"),
                         v.names = "reads", timevar = "stage", times = setdiff(names(mb_track), "sample"), idvar = "sample")
  long$stage <- factor(long$stage, levels = setdiff(names(mb_track), "sample"))
  save_fig(ggplot2::ggplot(long, ggplot2::aes(stage, reads, group = sample)) + ggplot2::geom_line(colour = "#9A8F84", alpha = 0.7) +
    ggplot2::geom_point(colour = "#C27840", size = 1) + ggplot2::labs(x = NULL, y = "reads per sample"), "n4__reads_through_the_steps")
  # The denoising objects hold every read's assignment and are not needed
  # after this step; kept, they cost PICRUSt2 the memory it needs later.
  rm(list = intersect(c("dd_f", "dd_r", "merged", "seqtab_all"), ls())); invisible(gc())
}
step_done("n4", 4, 4, "ASV Inference (DADA2)")
```

## Taxonomy Assignment (bio_expr)

Recorded figures from this step: `n5__classified_by_rank.png`.

```{r n5}
step_start("n5", 6, 15, "Taxonomy Assignment (bio_expr)")
settings <- list(`reference` = "SILVA 138.1 (16S)", `min_bootstrap` = 50, `species` = TRUE)   # this step's settings from the canvas (tested program, pinned by the platform)
# Taxonomy Assignment: the naive Bayesian classifier (the RDP method, as DADA2
# implements it) against a curated reference, and exact species matches for 16S.
if (!exists("ps") || is.null(ps)) {
  step_skipped("n5", "run ASV Inference (DADA2) or Import Feature Table first")
} else if (is.null(phyloseq::refseq(ps, errorIfNULL = FALSE))) {
  step_skipped("n5", "the feature table carries no ASV sequences to classify")
} else {
  refdir <- "/opt/alkhemyst/refs/microbiome"
  ref <- switch(settings$reference,
    "PR2 5.0.0 (18S)" = list(file = "pr2_version_5.0.0_SSU_dada2.fasta.gz", species = NULL, rc = FALSE,
      levels = c("Domain", "Supergroup", "Division", "Subdivision", "Class", "Order", "Family", "Genus", "Species")),
    "UNITE 10.0 (ITS, fungi)" = list(file = "unite_fungi_v10.0.fasta", species = NULL, rc = TRUE,
      levels = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")),
    list(file = "silva_nr99_v138.1_train_set.fa.gz", species = "silva_species_assignment_v138.1.fa.gz", rc = FALSE,
      levels = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus")))
  seqs <- as.character(phyloseq::refseq(ps))
  set.seed(100)
  taxa <- dada2::assignTaxonomy(seqs, file.path(refdir, ref$file), minBoot = as.integer(settings$min_bootstrap),
                                taxLevels = ref$levels, tryRC = ref$rc, multithread = 2, verbose = FALSE)
  if (!is.null(ref$species) && !isFALSE(settings$species)) {
    taxa <- dada2::addSpecies(taxa, file.path(refdir, ref$species), verbose = FALSE)
  }
  # UNITE labels ranks as "k__Fungi"; the prefix says nothing the column does not
  taxa[] <- sub("^[a-z]__", "", taxa)
  rownames(taxa) <- names(seqs)
  ps <- phyloseq::merge_phyloseq(ps, phyloseq::tax_table(taxa))
  manifest <- tryCatch(utils::read.delim(file.path(refdir, "MANIFEST.tsv")), error = function(e) NULL)
  used <- if (!is.null(manifest)) manifest$version[manifest$file == ref$file] else ref$file
  metric("n5", "reference", if (length(used)) used[1] else ref$file)
  rates <- data.frame(rank = colnames(taxa), classified_pct = round(100 * colMeans(!is.na(taxa)), 1))
  for (i in seq_len(nrow(rates))) metric("n5", paste0("classified_", tolower(rates$rank[i]), "_pct"), rates$classified_pct[i])
  emit_preview("n5", data.frame(asv = rownames(taxa), taxa, check.names = FALSE, row.names = NULL))
  data.table::fwrite(data.frame(asv = rownames(taxa), sequence = seqs, taxa, check.names = FALSE), "artifacts/n5__taxonomy.csv")
  rates$rank <- factor(rates$rank, levels = colnames(taxa))
  save_fig(ggplot2::ggplot(rates, ggplot2::aes(rank, classified_pct)) + ggplot2::geom_col(fill = "#C27840") +
    ggplot2::geom_text(ggplot2::aes(label = paste0(classified_pct, "%")), vjust = -0.3, size = 3) +
    ggplot2::labs(x = NULL, y = "ASVs classified (%)") + ggplot2::ylim(0, 105), "n5__classified_by_rank")
}
step_done("n5", 6, 15, "Taxonomy Assignment (bio_expr)")
```

## Phylogenetic Tree

Recorded figures from this step: `n6__tree.png`.

```{r n6}
step_start("n6", 6, 15, "Phylogenetic Tree")
settings <- list(`model` = "GTR")   # this step's settings from the canvas (tested program, pinned by the platform)
# Phylogenetic Tree: align the ASVs with MAFFT and infer an approximately
# maximum-likelihood tree with FastTree (GTR model), rooted at its midpoint.
if (!exists("ps") || is.null(ps) || is.null(phyloseq::refseq(ps, errorIfNULL = FALSE))) {
  step_skipped("n6", "run ASV Inference (DADA2) first: the tree is built from the ASV sequences")
} else if (phyloseq::ntaxa(ps) < 3) {
  step_skipped("n6", "fewer than three ASVs: there is no tree to build")
} else {
  fa <- tempfile(fileext = ".fasta"); aln <- tempfile(fileext = ".aln"); nwk <- tempfile(fileext = ".nwk")
  Biostrings::writeXStringSet(phyloseq::refseq(ps), fa)
  run_tool("mafft", c("--auto", "--thread", "2", "--quiet", fa), node_id = "n6", stdout_file = aln)
  run_tool("FastTree", c("-nt", if (identical(settings$model, "GTR")) "-gtr", "-gamma", "-quiet", aln), stdout_file = nwk)
  tree <- ape::read.tree(nwk)
  tree <- phangorn::midpoint(tree)
  ps <- phyloseq::merge_phyloseq(ps, phyloseq::phy_tree(tree))
  mb_tree <- tree
  ape::write.tree(tree, "artifacts/n6__asv_tree.nwk")
  metric("n6", "tips", ape::Ntip(tree))
  metric("n6", "alignment_columns", unique(Biostrings::width(Biostrings::readDNAStringSet(aln)))[1])
  tips <- data.frame(label = tree$tip.label)
  if (!is.null(phyloseq::tax_table(ps, errorIfNULL = FALSE))) {
    tt <- as.data.frame(as(phyloseq::tax_table(ps), "matrix"))
    tips$group <- ifelse(is.na(tt[tips$label, 2]), "Unclassified", tt[tips$label, 2])
  } else tips$group <- "ASV"
  p <- ggtree::`%<+%`(ggtree::ggtree(tree, layout = "circular", colour = "#9A8F84"), tips) +
    ggtree::geom_tippoint(ggplot2::aes(colour = group), size = 1.2) + ggplot2::labs(colour = NULL)
  save_fig(p, "n6__tree", width = 7.2, height = 6.4)
  emit_preview("n6", data.frame(asv = tree$tip.label, group = tips$group))
}
step_done("n6", 6, 15, "Phylogenetic Tree")
```

## BIOM Export (bio_expr)

Recorded figures from this step: `n7__reads_per_sample.png`.

```{r n7}
step_start("n7", 9, 15, "BIOM Export (bio_expr)")
settings <- list(`include_taxonomy` = TRUE, `include_samples` = TRUE)   # this step's settings from the canvas (tested program, pinned by the platform)
# BIOM Export: the feature table in the BIOM format every microbiome tool
# reads (QIIME 2, phyloseq, MicrobiomeAnalyst), with its taxonomy, sample
# table, sequences and tree beside it.
if (!exists("ps") || is.null(ps)) {
  step_skipped("n7", "there is no feature table yet: run ASV Inference (DADA2) or Import Feature Table first")
} else {
  otu <- as(phyloseq::otu_table(ps), "matrix")
  if (!phyloseq::taxa_are_rows(ps)) otu <- t(otu)
  tax <- if (!is.null(phyloseq::tax_table(ps, errorIfNULL = FALSE))) as.data.frame(as(phyloseq::tax_table(ps), "matrix")) else NULL
  sam <- if (!is.null(phyloseq::sample_data(ps, errorIfNULL = FALSE))) data.frame(phyloseq::sample_data(ps), check.names = FALSE) else NULL
  # BIOM 1.0 (JSON), written here rather than by biomformat::write_biom, which
  # puts each sample's metadata and each feature's taxonomy out as a bare array.
  # The Python biom reader behind QIIME 2 and PICRUSt2 refuses that ("Unable to
  # cast metadata"); the format wants a map per sample and {"taxonomy": [...]}.
  taxm <- if (!is.null(tax) && !isFALSE(settings$include_taxonomy)) as.matrix(tax) else NULL
  samm <- if (!is.null(sam) && !isFALSE(settings$include_samples)) sam else NULL
  nz <- which(otu != 0, arr.ind = TRUE)
  biom <- list(
    id = "n7 feature table", format = "Biological Observation Matrix 1.0.0", format_url = "http://biom-format.org",
    type = "OTU table", generated_by = "AlkhemYst-Ai", date = format(Sys.time(), "%Y-%m-%dT%H:%M:%S"),
    matrix_type = "sparse", matrix_element_type = if (all(otu == round(otu))) "int" else "float",
    shape = dim(otu), data = unname(cbind(nz[, 1] - 1, nz[, 2] - 1, otu[nz])),
    rows = lapply(rownames(otu), function(f) list(id = f, metadata = if (is.null(taxm)) NULL else
      list(taxonomy = I(ifelse(is.na(taxm[f, ]), "", unname(taxm[f, ])))))),
    columns = lapply(colnames(otu), function(s) list(id = s, metadata = if (is.null(samm)) NULL else
      lapply(as.list(samm[s, , drop = FALSE]), function(v) if (is.na(v)) NULL else as.character(v)))))
  jsonlite::write_json(biom, "artifacts/n7__feature_table.biom", auto_unbox = TRUE, null = "null", na = "null", digits = NA)
  data.table::fwrite(data.frame(feature = rownames(otu), otu, check.names = FALSE), "artifacts/n7__feature_table.csv")
  if (!is.null(tax)) data.table::fwrite(data.frame(feature = rownames(tax), tax, check.names = FALSE), "artifacts/n7__taxonomy.csv")
  if (!is.null(sam)) data.table::fwrite(sam, "artifacts/n7__sample_table.csv")
  if (!is.null(phyloseq::refseq(ps, errorIfNULL = FALSE))) Biostrings::writeXStringSet(phyloseq::refseq(ps), "artifacts/n7__sequences.fasta")
  if (!is.null(phyloseq::phy_tree(ps, errorIfNULL = FALSE))) ape::write.tree(phyloseq::phy_tree(ps), "artifacts/n7__tree.nwk")
  metric("n7", "features", nrow(otu))
  metric("n7", "samples", ncol(otu))
  metric("n7", "total_reads", sum(otu))
  emit_preview("n7", data.frame(feature = rownames(otu), otu[, seq_len(min(ncol(otu), 12)), drop = FALSE], check.names = FALSE))
  depth <- data.frame(sample = colnames(otu), reads = colSums(otu))
  save_fig(ggplot2::ggplot(depth, ggplot2::aes(stats::reorder(sample, reads), reads)) + ggplot2::geom_col(fill = "#C27840") +
    ggplot2::coord_flip() + ggplot2::labs(x = NULL, y = "reads in the table"), "n7__reads_per_sample")
}
step_done("n7", 9, 15, "BIOM Export (bio_expr)")
```

## Taxonomy Composition (bio_expr)

Recorded figures from this step: `n9__composition_by_group.png`, `n9__composition_by_sample.png`.

```{r n9}
step_start("n9", 10, 15, "Taxonomy Composition (bio_expr)")
settings <- list(`rank` = "Phylum", `top_n` = 10, `group_col` = "time")   # this step's settings from the canvas (tested program, pinned by the platform)
# Taxonomy Composition: relative abundance at a chosen rank, the most abundant
# taxa named and the rest pooled, per sample and per group.
if (!exists("ps") || is.null(ps) || is.null(phyloseq::tax_table(ps, errorIfNULL = FALSE))) {
  step_skipped("n9", "the feature table has no taxonomy: run Taxonomy Assignment first")
} else {
  rank <- settings$rank
  if (!(rank %in% phyloseq::rank_names(ps))) rank <- phyloseq::rank_names(ps)[min(2, length(phyloseq::rank_names(ps)))]
  otu <- as(phyloseq::otu_table(ps), "matrix"); if (phyloseq::taxa_are_rows(ps)) otu <- t(otu)
  tax <- as(phyloseq::tax_table(ps), "matrix")[colnames(otu), rank]
  tax[is.na(tax) | tax == ""] <- "Unclassified"
  agg <- t(rowsum(t(otu), tax))
  rel <- agg / pmax(rowSums(agg), 1)
  top <- names(sort(colMeans(rel), decreasing = TRUE))[seq_len(min(as.integer(settings$top_n), ncol(rel)))]
  other <- setdiff(colnames(rel), top)
  shown <- cbind(rel[, top, drop = FALSE], Other = if (length(other)) rowSums(rel[, other, drop = FALSE]) else 0)
  long <- data.frame(sample = rep(rownames(shown), ncol(shown)), taxon = rep(colnames(shown), each = nrow(shown)),
                     abundance = as.vector(shown))
  long$taxon <- factor(long$taxon, levels = c(top, "Other"))
  grp <- settings$group_col
  sam <- if (!is.null(phyloseq::sample_data(ps, errorIfNULL = FALSE))) data.frame(phyloseq::sample_data(ps), check.names = FALSE) else NULL
  has_grp <- !is.null(sam) && length(grp) && nzchar(grp) && grp %in% names(sam)
  pal <- grDevices::colorRampPalette(c("#C27840", "#E0B07A", "#6B8F71", "#3D5A80", "#98C1D9", "#9A8F84", "#A13D2D", "#EAD7C3"))(length(top))
  pal <- c(stats::setNames(pal, top), Other = "#D8D2CA")
  p <- ggplot2::ggplot(long, ggplot2::aes(sample, abundance, fill = taxon)) + ggplot2::geom_col(width = 0.9) +
    ggplot2::scale_fill_manual(values = pal) + ggplot2::scale_y_continuous(labels = function(x) paste0(100 * x, "%")) +
    ggplot2::labs(x = NULL, y = paste("relative abundance,", rank), fill = rank) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5))
  if (has_grp) {
    long$group <- as.character(sam[long$sample, grp]); long$group[is.na(long$group) | long$group == ""] <- "(none)"
    p <- p + ggplot2::facet_grid(~group, scales = "free_x", space = "free_x")
    # a sample with no group (a mock community, a blank) is drawn per sample
    # above and left out of the group means, where it would read as a group
    means <- stats::aggregate(abundance ~ group + taxon, data = long[long$group != "(none)", ], FUN = mean)
    save_fig(ggplot2::ggplot(means, ggplot2::aes(group, abundance, fill = taxon)) + ggplot2::geom_col() +
      ggplot2::scale_fill_manual(values = pal) + ggplot2::scale_y_continuous(labels = function(x) paste0(100 * x, "%")) +
      ggplot2::labs(x = grp, y = "mean relative abundance", fill = rank), "n9__composition_by_group")
  }
  save_fig(p, "n9__composition_by_sample", width = max(7.2, 0.28 * nrow(shown) + 3))
  metric("n9", "rank", rank)
  metric("n9", "taxa_at_rank", ncol(rel))
  metric("n9", "top_taxon", top[1])
  metric("n9", "top_taxon_mean_pct", round(100 * mean(rel[, top[1]]), 1))
  data.table::fwrite(data.frame(sample = rownames(rel), rel, check.names = FALSE), "artifacts/n9__relative_abundance.csv")
  emit_preview("n9", data.frame(taxon = colnames(shown), mean_pct = round(100 * colMeans(shown), 2)))
}
step_done("n9", 10, 15, "Taxonomy Composition (bio_expr)")
```

## Alpha Diversity

Recorded figures from this step: `n10__alpha_by_group.png`.

```{r n10}
step_start("n10", 11, 15, "Alpha Diversity")
settings <- list(`group_col` = "time", `rarefy` = FALSE, `rarefy_depth` = 0, `seed` = 1)   # this step's settings from the canvas (tested program, pinned by the platform)
# Alpha Diversity: richness and evenness within each sample, compared between groups.
if (!exists("ps") || is.null(ps)) {
  step_skipped("n10", "there is no feature table yet")
} else {
  x <- ps
  if (isTRUE(settings$rarefy)) {
    depth <- if (as.numeric(settings$rarefy_depth) > 0) as.numeric(settings$rarefy_depth) else min(phyloseq::sample_sums(x))
    x <- phyloseq::rarefy_even_depth(x, sample.size = depth, rngseed = as.integer(settings$seed), verbose = FALSE)
    metric("n10", "rarefied_to", depth)
  }
  est <- phyloseq::estimate_richness(x, measures = c("Observed", "Chao1", "Shannon", "Simpson"))
  rownames(est) <- phyloseq::sample_names(x)
  est$Pielou <- est$Shannon / log(pmax(est$Observed, 2))
  tree <- phyloseq::phy_tree(x, errorIfNULL = FALSE)
  if (!is.null(tree)) {
    otu <- as(phyloseq::otu_table(x), "matrix"); if (phyloseq::taxa_are_rows(x)) otu <- t(otu)
    est$Faith_PD <- picante::pd(otu, tree, include.root = FALSE)[rownames(est), "PD"]
  } else metric("n10", "faith_pd", "not computed: no phylogenetic tree (run Phylogenetic Tree first)")
  keep <- intersect(c("Observed", "Chao1", "Shannon", "Simpson", "Pielou", "Faith_PD"), names(est))
  est <- est[, keep, drop = FALSE]
  table_out <- data.frame(sample = rownames(est), round(est, 4), check.names = FALSE)
  grp <- settings$group_col
  sam <- if (!is.null(phyloseq::sample_data(x, errorIfNULL = FALSE))) data.frame(phyloseq::sample_data(x), check.names = FALSE) else NULL
  if (!is.null(sam) && length(grp) && nzchar(grp) && grp %in% names(sam)) {
    g <- as.character(sam[rownames(est), grp]); ok <- !is.na(g) & g != ""
    table_out[[grp]] <- g
    long <- data.frame(metric = rep(keep, each = sum(ok)), group = rep(g[ok], length(keep)),
                       value = unlist(lapply(keep, function(k) est[ok, k])))
    long$metric <- factor(long$metric, levels = keep)
    if (length(unique(g[ok])) >= 2) {
      for (k in keep) {
        kt <- stats::kruskal.test(est[ok, k], factor(g[ok]))
        metric("n10", paste0(k, "_p"), signif(kt$p.value, 3))
      }
      if (length(unique(g[ok])) > 2) {
        pw <- stats::pairwise.wilcox.test(est[ok, "Shannon"], g[ok], p.adjust.method = "BH")
        metric("n10", "Shannon_pairwise_min_p", signif(min(pw$p.value, na.rm = TRUE), 3))
      }
    }
    save_fig(ggplot2::ggplot(long, ggplot2::aes(group, value)) + ggplot2::geom_boxplot(outlier.shape = NA, fill = "#F3E6D8") +
      ggplot2::geom_jitter(width = 0.15, colour = "#C27840", size = 1.4) + ggplot2::facet_wrap(~metric, scales = "free_y") +
      ggplot2::labs(x = grp, y = NULL), "n10__alpha_by_group", height = 5.6)
  } else {
    long <- data.frame(metric = rep(keep, each = nrow(est)), value = unlist(est[keep]))
    save_fig(ggplot2::ggplot(long, ggplot2::aes(value)) + ggplot2::geom_histogram(bins = 15, fill = "#C27840") +
      ggplot2::facet_wrap(~metric, scales = "free"), "n10__alpha_distributions")
  }
  for (k in keep) metric("n10", paste0(k, "_median"), round(stats::median(est[[k]]), 3))
  emit_preview("n10", table_out)
  data.table::fwrite(table_out, "artifacts/n10__alpha_diversity.csv")
}
step_done("n10", 11, 15, "Alpha Diversity")
```

## Beta Diversity

Recorded figures from this step: `n11__pcoa.png`.

```{r n11}
step_start("n11", 12, 15, "Beta Diversity")
settings <- list(`distances` = c("Bray-Curtis", "Jaccard", "Unweighted UniFrac", "Weighted UniFrac"), `group_col` = "time", `permutations` = 999, `seed` = 1)   # this step's settings from the canvas (tested program, pinned by the platform)
# Beta Diversity: how different the samples are from each other, drawn as a
# PCoA for each distance and tested between groups with PERMANOVA.
if (!exists("ps") || is.null(ps)) {
  step_skipped("n11", "there is no feature table yet")
} else {
  x <- phyloseq::prune_samples(phyloseq::sample_sums(ps) > 0, ps)
  grp <- settings$group_col
  sam <- if (!is.null(phyloseq::sample_data(x, errorIfNULL = FALSE))) data.frame(phyloseq::sample_data(x), check.names = FALSE) else NULL
  has_grp <- !is.null(sam) && length(grp) && nzchar(grp) && grp %in% names(sam)
  # Compared by group, the ordination shows the samples the test compares. A
  # sample with no group (a mock community, a blank) is left out, and named:
  # kept, the MiSeq SOP mock took the whole first axis of the Bray-Curtis PCoA.
  if (has_grp) {
    gv <- as.character(sam[phyloseq::sample_names(x), grp])
    out <- phyloseq::sample_names(x)[is.na(gv) | gv == ""]
    if (length(out) && length(out) < phyloseq::nsamples(x) - 3) {
      x <- phyloseq::prune_samples(setdiff(phyloseq::sample_names(x), out), x)
      metric("n11", "samples_without_group_left_out", paste(out, collapse = ", "))
    }
  }
  rel <- phyloseq::transform_sample_counts(x, function(v) v / sum(v))
  otu <- as(phyloseq::otu_table(rel), "matrix"); if (phyloseq::taxa_are_rows(rel)) otu <- t(otu)
  tree <- phyloseq::phy_tree(x, errorIfNULL = FALSE)
  wanted <- trimws(unlist(strsplit(as.character(unlist(settings$distances)), ",")))
  dists <- list()
  if ("Bray-Curtis" %in% wanted) dists[["Bray-Curtis"]] <- vegan::vegdist(otu, "bray")
  if ("Jaccard" %in% wanted) dists[["Jaccard"]] <- vegan::vegdist(otu > 0, "jaccard", binary = TRUE)
  if (!is.null(tree) && "Unweighted UniFrac" %in% wanted) dists[["Unweighted UniFrac"]] <- phyloseq::UniFrac(x, weighted = FALSE)
  if (!is.null(tree) && "Weighted UniFrac" %in% wanted) dists[["Weighted UniFrac"]] <- phyloseq::UniFrac(x, weighted = TRUE)
  if (is.null(tree) && any(grepl("UniFrac", wanted))) metric("n11", "unifrac", "not computed: no phylogenetic tree (run Phylogenetic Tree first)")
  pts <- list(); tests <- list()
  for (nm in names(dists)) {
    d <- dists[[nm]]
    pc <- stats::cmdscale(d, k = 2, eig = TRUE)
    ev <- pc$eig[pc$eig > 0]; pct <- round(100 * pc$eig[1:2] / sum(ev), 1)
    pts[[nm]] <- data.frame(sample = rownames(pc$points), PC1 = pc$points[, 1], PC2 = pc$points[, 2],
                            distance = sprintf("%s (PC1 %s%%, PC2 %s%%)", nm, pct[1], pct[2]))
    if (has_grp) {
      g <- as.character(sam[labels(d), grp]); ok <- !is.na(g) & g != ""
      if (length(unique(g[ok])) >= 2 && sum(ok) >= 4) {
        dd <- stats::as.dist(as.matrix(d)[ok, ok])
        md <- data.frame(group = factor(g[ok]))
        set.seed(as.integer(settings$seed))
        a <- vegan::adonis2(dd ~ group, data = md, permutations = as.integer(settings$permutations))
        disp <- vegan::betadisper(dd, md$group)
        bd <- vegan::permutest(disp, permutations = as.integer(settings$permutations))
        # which group varies more: its samples' mean distance to the group centroid
        spread <- tapply(disp$distances, disp$group, mean)
        spread_txt <- paste(sprintf("%s %.3f", names(spread), spread), collapse = "; ")
        tests[[nm]] <- data.frame(distance = nm, R2 = round(a$R2[1], 4), F = round(a$F[1], 3), p = a$`Pr(>F)`[1],
                                  dispersion_p = bd$tab$`Pr(>F)`[1], spread_by_group = spread_txt)
        key <- gsub("[^a-z]+", "_", tolower(nm))
        metric("n11", paste0(key, "_permanova_R2"), round(a$R2[1], 4))
        metric("n11", paste0(key, "_permanova_p"), a$`Pr(>F)`[1])
        metric("n11", paste0(key, "_dispersion_p"), bd$tab$`Pr(>F)`[1])
        metric("n11", paste0(key, "_spread_by_group"), spread_txt)
      }
    }
  }
  allp <- do.call(rbind, pts)
  if (has_grp) { allp$group <- as.character(sam[allp$sample, grp]); allp$group[is.na(allp$group) | allp$group == ""] <- "(none)" } else allp$group <- "all"
  p <- ggplot2::ggplot(allp, ggplot2::aes(PC1, PC2, colour = group)) + ggplot2::geom_point(size = 2) +
    ggplot2::facet_wrap(~distance, scales = "free") + ggplot2::labs(colour = if (has_grp) grp else NULL)
  if (has_grp) p <- p + ggplot2::stat_ellipse(data = allp[allp$group != "(none)", ], level = 0.95, linetype = "dashed", show.legend = FALSE)
  save_fig(p, "n11__pcoa", width = 8.4, height = if (length(dists) > 2) 7 else 4.2)
  if (length(tests)) {
    tt <- do.call(rbind, tests); emit_preview("n11", tt)
    data.table::fwrite(tt, "artifacts/n11__permanova.csv")
  } else emit_preview("n11", allp)
  metric("n11", "distances", paste(names(dists), collapse = ", "))
}
step_done("n11", 12, 15, "Beta Diversity")
```

## Differential Abundance (Microbiome) (bio_expr)

Recorded figures from this step: `n12__differential_taxa.png`, `n12__volcano.png`.

```{r n12}
step_start("n12", 13, 15, "Differential Abundance (Microbiome) (bio_expr)")
settings <- list(`method` = "ANCOM-BC2", `group_col` = "time", `reference_level` = "Early", `compare_level` = "Late", `rank` = "Genus", `min_prevalence` = 0.1, `covariates` = character(0), `alpha` = 0.05, `lda_threshold` = 2, `seed` = 1, `exclude_columns` = character(0))   # this step's settings from the canvas (tested program, pinned by the platform)
# Differential Abundance (microbiome): which taxa differ between two groups,
# by a method built for compositional count data. Each method answers the
# same question differently; the result table has the same columns for all.
if (!exists("ps") || is.null(ps) || is.null(phyloseq::sample_data(ps, errorIfNULL = FALSE))) {
  step_skipped("n12", "the feature table has no sample table with the groups to compare")
} else {
  grp <- settings$group_col
  sam <- data.frame(phyloseq::sample_data(ps), check.names = FALSE)
  if (!length(grp) || !nzchar(grp) || !(grp %in% names(sam))) {
    step_skipped("n12", paste0("the group column '", grp, "' is not in the sample table"))
  } else {
    x <- ps
    rank <- settings$rank
    if (length(rank) && nzchar(rank) && rank != "ASV" && !is.null(phyloseq::tax_table(x, errorIfNULL = FALSE)) && rank %in% phyloseq::rank_names(x)) {
      x <- phyloseq::tax_glom(x, rank, NArm = FALSE)
    }
    g <- as.character(sam[phyloseq::sample_names(x), grp])
    x <- phyloseq::prune_samples(!is.na(g) & g != "", x)
    g <- as.character(data.frame(phyloseq::sample_data(x), check.names = FALSE)[[grp]])
    levels_all <- sort(unique(g))
    ref <- if (length(settings$reference_level) && settings$reference_level %in% levels_all) settings$reference_level else levels_all[1]
    test <- if (length(settings$compare_level) && settings$compare_level %in% setdiff(levels_all, ref)) settings$compare_level else setdiff(levels_all, ref)[1]
    x <- phyloseq::prune_samples(g %in% c(ref, test), x)
    counts <- as(phyloseq::otu_table(x), "matrix"); if (!phyloseq::taxa_are_rows(x)) counts <- t(counts)
    prev <- rowMeans(counts > 0)
    counts <- counts[prev >= as.numeric(settings$min_prevalence) & rowSums(counts) > 0, , drop = FALSE]
    x <- phyloseq::prune_taxa(rownames(counts), x)
    md <- data.frame(phyloseq::sample_data(x), check.names = FALSE)
    md$group <- factor(as.character(md[[grp]]), levels = c(ref, test))
    covs <- intersect(trimws(unlist(strsplit(as.character(unlist(settings$covariates)), ","))), names(md))   # a list of columns, or "a, b"
    rhs <- paste(c("group", covs), collapse = " + ")
    alpha <- as.numeric(settings$alpha)
    label <- rownames(counts)
    if (!is.null(phyloseq::tax_table(x, errorIfNULL = FALSE))) {
      tt <- as(phyloseq::tax_table(x), "matrix")[rownames(counts), , drop = FALSE]
      best <- apply(tt, 1, function(r) { r <- r[!is.na(r) & r != ""]; if (length(r)) utils::tail(r, 1) else NA })
      label <- ifelse(is.na(best), rownames(counts), paste0(best, " (", rownames(counts), ")"))
    }
    method <- settings$method
    res <- NULL
    if (nrow(counts) < 2 || min(table(md$group)) < 2) {
      step_skipped("n12", sprintf("too little to compare: %d features and %s", nrow(counts), paste(names(table(md$group)), table(md$group), sep = " n=", collapse = ", ")))
    } else if (method == "ALDEx2") {
      set.seed(as.integer(settings$seed))
      a <- ALDEx2::aldex(as.data.frame(counts), as.character(md$group), mc.samples = 128, test = "t", effect = TRUE, denom = "all", verbose = FALSE)
      res <- data.frame(feature = rownames(a), effect = a$effect, p = a$we.ep, q = a$we.eBH)
    } else if (method == "DESeq2") {
      dds <- DESeq2::DESeqDataSetFromMatrix(countData = round(counts), colData = md, design = stats::as.formula(paste("~", rhs)))
      dds <- DESeq2::DESeq(dds, sfType = "poscounts", fitType = "local", quiet = TRUE)
      r <- DESeq2::results(dds, contrast = c("group", test, ref))
      res <- data.frame(feature = rownames(r), effect = r$log2FoldChange, p = r$pvalue, q = r$padj)
    } else if (method == "ANCOM-BC2") {
      phyloseq::sample_data(x) <- phyloseq::sample_data(md)
      o <- ANCOMBC::ancombc2(data = x, fix_formula = rhs, p_adj_method = "holm", prv_cut = 0, group = "group",
                             struc_zero = FALSE, neg_lb = FALSE, alpha = alpha, global = FALSE, n_cl = 1, verbose = FALSE)$res
      col <- paste0("group", test)
      res <- data.frame(feature = o$taxon, effect = o[[paste0("lfc_", col)]], p = o[[paste0("p_", col)]], q = o[[paste0("q_", col)]])
    } else if (method == "LinDA") {
      o <- MicrobiomeStat::linda(feature.dat = counts, meta.dat = md, formula = paste("~", rhs), feature.dat.type = "count",
                                 prev.filter = 0, alpha = alpha, verbose = FALSE)$output[[paste0("group", test)]]
      res <- data.frame(feature = rownames(o), effect = o$log2FoldChange, p = o$pvalue, q = o$padj)
    } else if (method == "MaAsLin2") {
      md2 <- md; md2$group <- as.character(md2$group)
      o <- Maaslin2::Maaslin2(input_data = as.data.frame(t(counts)), input_metadata = md2, output = tempfile("maaslin2_"),
                              fixed_effects = c("group", covs), reference = paste0("group,", ref), normalization = "TSS",
                              transform = "LOG", analysis_method = "LM", min_prevalence = 0, plot_heatmap = FALSE,
                              plot_scatter = FALSE, cores = 1)$results
      o <- o[o$metadata == "group", ]
      res <- data.frame(feature = o$feature, effect = o$coef, p = o$pval, q = o$qval)
    } else if (method == "LEfSe") {
      rel <- sweep(counts, 2, pmax(colSums(counts), 1), "/") * 1e6
      se <- SummarizedExperiment::SummarizedExperiment(assays = list(counts = rel), colData = md)
      f <- names(formals(lefser::lefser))
      col_arg <- intersect(c("groupCol", "classCol", "grp"), f)[1]
      args <- list(se, kruskal.threshold = alpha, lda.threshold = as.numeric(settings$lda_threshold))
      args[[col_arg]] <- "group"
      o <- do.call(lefser::lefser, args)
      o <- as.data.frame(o)
      res <- data.frame(feature = o[[1]], effect = o$scores, p = NA_real_, q = NA_real_)
    }
    if (!is.null(res)) {
      res$taxon <- label[match(res$feature, rownames(counts))]
      res$significant <- if (method == "LEfSe") TRUE else !is.na(res$q) & res$q < alpha
      res <- res[order(res$q, -abs(res$effect)), c("feature", "taxon", "effect", "p", "q", "significant")]
      metric("n12", "method", method)
      metric("n12", "comparison", paste(test, "vs", ref))
      metric("n12", "features_tested", nrow(counts))
      metric("n12", "significant", sum(res$significant, na.rm = TRUE))
      emit_preview("n12", res)
      data.table::fwrite(res, "artifacts/n12__differential_abundance.csv")
      sig <- res[res$significant & !is.na(res$effect), ]
      top <- utils::head(sig[order(-abs(sig$effect)), ], 25)
      if (nrow(top)) {
        top$taxon <- factor(top$taxon, levels = top$taxon[order(top$effect)])
        top$direction <- ifelse(top$effect > 0, paste("higher in", test), paste("higher in", ref))
        save_fig(ggplot2::ggplot(top, ggplot2::aes(effect, taxon, fill = direction)) + ggplot2::geom_col() +
          ggplot2::scale_fill_manual(values = stats::setNames(c("#C27840", "#3D5A80"), c(paste("higher in", test), paste("higher in", ref)))) +
          ggplot2::labs(x = switch(method, ALDEx2 = "effect size", LEfSe = "LDA score (log10)", "log fold change"), y = NULL, fill = NULL),
          "n12__differential_taxa")
      }
      if (method != "LEfSe") {
        save_fig(ggplot2::ggplot(res, ggplot2::aes(effect, -log10(pmax(q, 1e-300)), colour = significant)) + ggplot2::geom_point(alpha = 0.8) +
          ggplot2::scale_colour_manual(values = c(`TRUE` = "#C27840", `FALSE` = "#9A8F84")) +
          ggplot2::geom_hline(yintercept = -log10(alpha), linetype = "dashed") +
          ggplot2::labs(x = "effect", y = "-log10 adjusted p", colour = paste("q <", alpha)), "n12__volcano")
      }
    }
  }
}
# the fitted models (ALDEx2 Monte Carlo instances, DESeq2) are written out above and not used again
rm(list = intersect(c("x", "dds", "se"), ls())); invisible(gc())
step_done("n12", 13, 15, "Differential Abundance (Microbiome) (bio_expr)")
```

## Functional Prediction (PICRUSt2)

Recorded figures from this step: `n13__nsti.png`, `n13__top_metacyc_pathways.png`.

```{r n13}
step_start("n13", 6, 15, "Functional Prediction (PICRUSt2)")
settings <- list(`max_nsti` = 2, `kegg_pathways` = TRUE, `stratified` = FALSE)   # this step's settings from the canvas (tested program, pinned by the platform)
# Functional Prediction (PICRUSt2): place each ASV in a reference tree of
# sequenced genomes and predict the gene families (KEGG orthologs, EC numbers)
# and MetaCyc pathways each sample's community carries. A prediction from the
# marker gene, not a measurement: NSTI says how far each ASV is from a genome.
if (!exists("ps") || is.null(ps) || is.null(phyloseq::refseq(ps, errorIfNULL = FALSE))) {
  step_skipped("n13", "PICRUSt2 needs the ASV sequences: run ASV Inference (DADA2) first")
} else {
  counts <- as(phyloseq::otu_table(ps), "matrix"); if (!phyloseq::taxa_are_rows(ps)) counts <- t(counts)
  fa <- tempfile(fileext = ".fasta"); tab <- tempfile(fileext = ".tsv"); out <- file.path(tempdir(), "picrust2_out")
  Biostrings::writeXStringSet(phyloseq::refseq(ps), fa)
  data.table::fwrite(data.frame(`#OTU ID` = rownames(counts), counts, check.names = FALSE), tab, sep = "\t")
  # Placement with SEPP, not PICRUSt2's default EPA-ng: EPA-ng holds the whole
  # 20,000-genome reference in memory and needed more than 6.8 GB on this
  # runner (killed by the kernel), where SEPP places into subsets of the tree
  # and the whole pipeline peaked at 3.8 GB. Both are PICRUSt2's own methods.
  invisible(gc())
  # A prediction that cannot finish is reported and the run goes on: the steps
  # after it do not need it, and a killed PICRUSt2 must not take them down.
  ran <- tryCatch({
    run_tool("picrust2_pipeline.py", c("-s", fa, "-i", tab, "-o", out, "-p", "2", "-t", "sepp", "--max_nsti", as.character(settings$max_nsti),
                                       if (isTRUE(settings$stratified)) "--stratified"), node_id = "n13", timeout = 14400)
    TRUE
  }, error = function(e) conditionMessage(e))
  if (!isTRUE(ran)) {
    killed <- grepl("status -9|status 137|status -15", ran)
    step_skipped("n13", paste0("PICRUSt2 did not finish",
      if (killed) " (stopped by the system, most often for lack of memory; fewer ASVs or a larger runner)" else "",
      ": ", substr(ran, 1, 400)))
  } else {
  rd <- function(f) data.table::fread(f, data.table = FALSE)
  mapdir <- "/opt/picrust2/lib/python3.9/site-packages/picrust2/default_files"
  picrust <- list(
    ko = rd(file.path(out, "KO_metagenome_out", "pred_metagenome_unstrat.tsv.gz")),
    ec = rd(file.path(out, "EC_metagenome_out", "pred_metagenome_unstrat.tsv.gz")),
    metacyc = rd(file.path(out, "pathways_out", "path_abun_unstrat.tsv.gz")))
  if (!isFALSE(settings$kegg_pathways)) {
    kout <- file.path(tempdir(), "picrust2_kegg")
    kegg <- tryCatch({
      run_tool("pathway_pipeline.py", c("-i", file.path(out, "KO_metagenome_out", "pred_metagenome_unstrat.tsv.gz"), "-o", kout,
                                        "--no_regroup", "--map", file.path(mapdir, "pathway_mapfiles", "KEGG_pathways_to_KO.tsv"), "-p", "2"),
               node_id = "n13")
      rd(file.path(kout, "path_abun_unstrat.tsv.gz"))
    }, error = function(e) { metric("n13", "kegg_pathways_not_computed", substr(conditionMessage(e), 1, 300)); NULL })
    if (!is.null(kegg)) picrust$kegg <- kegg
  }
  nsti <- rd(file.path(out, "marker_predicted_and_nsti.tsv.gz"))
  names(nsti)[1] <- "asv"
  kept <- nsti$metadata_NSTI <= as.numeric(settings$max_nsti)
  w <- rowSums(counts)[nsti$asv]
  metric("n13", "asvs", nrow(counts))
  metric("n13", "asvs_excluded_by_nsti", sum(!kept))
  metric("n13", "weighted_mean_nsti", round(stats::weighted.mean(nsti$metadata_NSTI, w, na.rm = TRUE), 3))
  metric("n13", "ko_families", nrow(picrust$ko))
  metric("n13", "ec_numbers", nrow(picrust$ec))
  metric("n13", "metacyc_pathways", nrow(picrust$metacyc))
  if (!is.null(picrust$kegg)) metric("n13", "kegg_pathways", nrow(picrust$kegg))
  for (nm in names(picrust)) data.table::fwrite(picrust[[nm]], sprintf("artifacts/n13__%s_abundance.csv", nm))
  desc <- rd(file.path(mapdir, "description_mapfiles", "metacyc_pathways_info.txt.gz"))
  names(desc) <- c("pathway", "description")
  m <- as.matrix(picrust$metacyc[, -1]); rownames(m) <- picrust$metacyc[[1]]
  rel <- sweep(m, 2, pmax(colSums(m), 1), "/")
  top <- names(sort(rowMeans(rel), decreasing = TRUE))[seq_len(min(20, nrow(rel)))]
  z <- t(scale(t(rel[top, , drop = FALSE]))); z[is.na(z)] <- 0
  lab <- desc$description[match(top, desc$pathway)]; lab <- ifelse(is.na(lab), top, lab)
  hm <- data.frame(pathway = rep(lab, ncol(z)), sample = rep(colnames(z), each = nrow(z)), z = as.vector(z))
  save_fig(ggplot2::ggplot(hm, ggplot2::aes(sample, pathway, fill = z)) + ggplot2::geom_tile() +
    ggplot2::scale_fill_gradient2(low = "#3D5A80", mid = "#FBF9F5", high = "#C27840") +
    ggplot2::labs(x = NULL, y = NULL, fill = "z-score") +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5)), "n13__top_metacyc_pathways",
    width = 9, height = 6.5)
  save_fig(ggplot2::ggplot(nsti, ggplot2::aes(metadata_NSTI)) + ggplot2::geom_histogram(bins = 30, fill = "#C27840") +
    ggplot2::geom_vline(xintercept = as.numeric(settings$max_nsti), linetype = "dashed") +
    ggplot2::labs(x = "NSTI (distance to the nearest sequenced genome)", y = "ASVs"), "n13__nsti")
  emit_preview("n13", data.frame(pathway = top, description = lab, mean_relative_abundance = round(rowMeans(rel[top, , drop = FALSE]), 5)))
}
}
step_done("n13", 6, 15, "Functional Prediction (PICRUSt2)")
```

## Pathway Abundance

Recorded figures from this step: `n14__pathway_heatmap.png`.

```{r n14}
step_start("n14", 7, 15, "Pathway Abundance")
settings <- list(`source` = "MetaCyc pathways", `group_col` = "time", `reference_level` = "Early", `compare_level` = "Late", `method` = "ALDEx2", `alpha` = 0.05, `top_n` = 20)   # this step's settings from the canvas (tested program, pinned by the platform)
# Pathway Abundance: compare the predicted functions between groups, as
# MetaCyc or KEGG pathways, KEGG orthologs or EC numbers, with their names.
if (!exists("picrust")) {
  step_skipped("n14", "run Functional Prediction (PICRUSt2) first")
} else if (!exists("ps") || is.null(ps) || is.null(phyloseq::sample_data(ps, errorIfNULL = FALSE))) {
  step_skipped("n14", "there is no sample table with the groups to compare")
} else {
  mapdir <- "/opt/picrust2/lib/python3.9/site-packages/picrust2/default_files/description_mapfiles"
  src <- settings$source
  tab <- switch(src, "KEGG pathways" = picrust$kegg, "KEGG orthologs" = picrust$ko, "EC numbers" = picrust$ec, picrust$metacyc)
  descfile <- switch(src, "KEGG pathways" = "KEGG_pathways_info.tsv.gz", "KEGG orthologs" = "ko_info.tsv.gz",
                     "EC numbers" = "ec_level4_info.tsv.gz", "metacyc_pathways_info.txt.gz")
  grp <- settings$group_col
  sam <- data.frame(phyloseq::sample_data(ps), check.names = FALSE)
  if (is.null(tab)) {
    step_skipped("n14", paste(src, "were not predicted: turn on KEGG pathways in Functional Prediction"))
  } else if (!length(grp) || !nzchar(grp) || !(grp %in% names(sam))) {
    step_skipped("n14", paste0("the group column '", grp, "' is not in the sample table"))
  } else {
    m <- as.matrix(tab[, -1]); rownames(m) <- tab[[1]]
    m <- m[, intersect(colnames(m), rownames(sam)), drop = FALSE]
    g <- as.character(sam[colnames(m), grp]); ok <- !is.na(g) & g != ""
    m <- m[, ok, drop = FALSE]; g <- g[ok]
    lv <- sort(unique(g))
    ref <- if (length(settings$reference_level) && settings$reference_level %in% lv) settings$reference_level else lv[1]
    test <- if (length(settings$compare_level) && settings$compare_level %in% setdiff(lv, ref)) settings$compare_level else setdiff(lv, ref)[1]
    m <- m[, g %in% c(ref, test), drop = FALSE]; g <- g[g %in% c(ref, test)]
    m <- m[rowSums(m > 0) >= 2, , drop = FALSE]
    desc <- data.table::fread(file.path(mapdir, descfile), header = FALSE, data.table = FALSE)
    lab <- desc[[2]][match(rownames(m), desc[[1]])]; lab <- ifelse(is.na(lab), rownames(m), lab)
    alpha <- as.numeric(settings$alpha)
    rel <- sweep(m, 2, pmax(colSums(m), 1), "/")
    if (settings$method == "ALDEx2") {
      set.seed(1)
      a <- ALDEx2::aldex(as.data.frame(round(m)), g, mc.samples = 128, test = "t", effect = TRUE, denom = "all", verbose = FALSE)
      res <- data.frame(feature = rownames(a), effect = a$effect, p = a$we.ep, q = a$we.eBH)
    } else if (settings$method == "LinDA") {
      md <- data.frame(group = factor(g, levels = c(ref, test)), row.names = colnames(m))
      o <- MicrobiomeStat::linda(feature.dat = round(m), meta.dat = md, formula = "~ group", feature.dat.type = "count",
                                 prev.filter = 0, alpha = alpha, verbose = FALSE)$output[[paste0("group", test)]]
      res <- data.frame(feature = rownames(o), effect = o$log2FoldChange, p = o$pvalue, q = o$padj)
    } else {
      p <- apply(rel, 1, function(v) stats::wilcox.test(v[g == test], v[g == ref], exact = FALSE)$p.value)
      eff <- log2((rowMeans(rel[, g == test, drop = FALSE]) + 1e-9) / (rowMeans(rel[, g == ref, drop = FALSE]) + 1e-9))
      res <- data.frame(feature = rownames(rel), effect = eff, p = p, q = stats::p.adjust(p, "BH"))
    }
    res$description <- lab[match(res$feature, rownames(m))]
    res$significant <- !is.na(res$q) & res$q < alpha
    res <- res[order(res$q, -abs(res$effect)), c("feature", "description", "effect", "p", "q", "significant")]
    metric("n14", "source", src)
    metric("n14", "method", settings$method)
    metric("n14", "comparison", paste(test, "vs", ref))
    metric("n14", "tested", nrow(res))
    metric("n14", "significant", sum(res$significant))
    emit_preview("n14", res)
    data.table::fwrite(res, "artifacts/n14__pathway_differences.csv")
    show <- utils::head(if (sum(res$significant)) res[res$significant, ] else res, as.integer(settings$top_n))
    z <- t(scale(t(rel[show$feature, , drop = FALSE]))); z[is.na(z)] <- 0
    hm <- data.frame(feature = rep(show$description, ncol(z)), sample = rep(colnames(z), each = nrow(z)),
                     group = rep(g, each = nrow(z)), z = as.vector(z))
    hm$feature <- factor(hm$feature, levels = rev(unique(show$description)))
    save_fig(ggplot2::ggplot(hm, ggplot2::aes(sample, feature, fill = z)) + ggplot2::geom_tile() +
      ggplot2::facet_grid(~group, scales = "free_x", space = "free_x") +
      ggplot2::scale_fill_gradient2(low = "#3D5A80", mid = "#FBF9F5", high = "#C27840") +
      ggplot2::labs(x = NULL, y = NULL, fill = "z-score", title = if (sum(res$significant)) paste0("q < ", alpha) else "none significant; the most different shown") +
      ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5)), "n14__pathway_heatmap",
      width = 9, height = max(4, 0.3 * nrow(show) + 2))
  }
}
step_done("n14", 7, 15, "Pathway Abundance")
```

## Mock Community Check

Recorded figures from this step: `n15__mock_community.png`.

```{r n15}
step_start("n15", 5, 15, "Mock Community Check")
settings <- list(`mock_sample` = "Mock", `reference_file` = "HMP_MOCK.v35.fasta", `min_reads` = 1)   # this step's settings from the canvas (tested program, pinned by the platform)
# Mock Community Check: a sample of known composition, sequenced with the study,
# is the one place where the right answer is known. The ASVs found in it are
# compared with the expected sequences: how many expected strains were
# recovered exactly, how many ASVs match nothing expected, and how much of the
# sample those unexpected ASVs hold (contamination, chimeras, errors left in).
if (!exists("ps") || is.null(ps) || is.null(phyloseq::refseq(ps, errorIfNULL = FALSE))) {
  step_skipped("n15", "the check needs the ASV sequences: run ASV Inference (DADA2), or import a table with its sequences")
} else {
  samples <- phyloseq::sample_names(ps)
  mock <- settings$mock_sample
  if (!length(mock) || !nzchar(mock)) mock <- samples[grepl("mock", samples, ignore.case = TRUE)][1]
  fastas <- list.files("data", pattern = "[.](fasta|fa|fna)(.gz)?$", full.names = TRUE, ignore.case = TRUE)
  refname <- settings$reference_file
  ref_path <- if (length(refname) && nzchar(refname)) file.path("data", basename(refname)) else
    fastas[grepl("mock", basename(fastas), ignore.case = TRUE)][1]
  if (is.na(mock) || !(mock %in% samples)) {
    step_skipped("n15", sprintf("no mock sample: name it in the settings (samples: %s)", paste(utils::head(samples, 8), collapse = ", ")))
  } else if (!length(ref_path) || is.na(ref_path) || !file.exists(ref_path)) {
    step_skipped("n15", "no reference of the expected sequences: add a FASTA of the mock's strains to the dataset and name it in the settings")
  } else {
    ref <- Biostrings::readDNAStringSet(ref_path)
    counts <- as(phyloseq::otu_table(ps), "matrix"); if (!phyloseq::taxa_are_rows(ps)) counts <- t(counts)
    in_mock <- counts[, mock]
    in_mock <- sort(in_mock[in_mock >= max(1, as.numeric(settings$min_reads))], decreasing = TRUE)
    seqs <- phyloseq::refseq(ps)[names(in_mock)]
    both <- c(ref, Biostrings::reverseComplement(ref))
    # the fewest mismatches against any expected sequence (an ASV is a stretch of
    # the full-length 16S reference, so it is searched for inside it)
    nearest <- vapply(as.character(seqs), function(s) {
      for (k in 0:3) if (sum(Biostrings::vcountPattern(s, both, max.mismatch = k)) > 0) return(k)
      NA_real_
    }, numeric(1))
    # A strain can carry several 16S copies ("S.aureus.1", "S.aureus.2"): strains
    # are counted by name, sequences by entry. Strains whose region is identical
    # share one ASV, so an ASV names every strain it matches exactly.
    strain_of <- function(x) sub("[._-][0-9]+$", "", sub("\\s.*", "", x))
    hit_ref <- vapply(as.character(seqs), function(s) {
      i <- which(Biostrings::vcountPattern(s, both, max.mismatch = 0) > 0)
      if (length(i)) paste(unique(strain_of(names(both)[i])), collapse = "; ") else NA_character_
    }, character(1))
    res <- data.frame(asv = names(in_mock), reads = as.numeric(in_mock),
                      share = round(100 * as.numeric(in_mock) / sum(in_mock), 2),
                      mismatches = nearest, expected_strain = unname(hit_ref),
                      verdict = ifelse(!is.na(nearest) & nearest == 0, "exact match",
                                ifelse(!is.na(nearest), sprintf("%d mismatch%s", nearest, ifelse(nearest == 1, "", "es")), "not expected")))
    exact <- res$verdict == "exact match"
    strains <- unique(strain_of(names(ref)))
    found_strains <- unique(unlist(strsplit(stats::na.omit(res$expected_strain), "; ")))
    recovered <- length(found_strains)
    expected <- length(strains)
    metric("n15", "mock_sample", mock)
    metric("n15", "expected_strains", expected)
    metric("n15", "expected_sequences", length(ref))
    metric("n15", "asvs_shared_by_strains", sum(grepl(";", res$expected_strain)))
    metric("n15", "strains_missing", if (length(setdiff(strains, found_strains))) paste(setdiff(strains, found_strains), collapse = ", ") else "none")
    metric("n15", "asvs_in_mock", nrow(res))
    metric("n15", "asvs_exact_match", sum(exact))
    metric("n15", "strains_recovered", recovered)
    metric("n15", "asvs_not_expected", sum(res$verdict == "not expected"))
    metric("n15", "reads_not_exact_pct", round(100 * sum(res$reads[!exact]) / max(sum(res$reads), 1), 2))
    data.table::fwrite(res, "artifacts/n15__mock_community.csv")
    emit_preview("n15", res)
    res$asv <- factor(res$asv, levels = rev(res$asv))
    save_fig(ggplot2::ggplot(res, ggplot2::aes(asv, reads, fill = verdict)) + ggplot2::geom_col() + ggplot2::coord_flip() +
      ggplot2::scale_fill_manual(values = c("exact match" = "#C27840", "1 mismatch" = "#D9A577", "2 mismatches" = "#D9A577",
                                            "3 mismatches" = "#D9A577", "not expected" = "#3D5A80")) +
      ggplot2::labs(x = NULL, y = sprintf("reads in %s", mock), fill = NULL,
                    title = sprintf("%d of %d expected strains recovered exactly; %d ASV(s) not expected",
                                    recovered, expected, sum(res$verdict == "not expected"))),
      "n15__mock_community", width = 7.2, height = max(4, 0.25 * nrow(res) + 1.5))
  }
}
step_done("n15", 5, 15, "Mock Community Check")
```

## Finalize

```{r step-2}
local({
  out <- as.list(METRICS)
  if (length(out) == 0) {
    # Said plainly rather than written as an empty object. A run that measured
    # nothing is nearly always a run where the steps wrote somewhere else, and
    # an empty metrics.json looks identical to a pipeline that had nothing to
    # measure.
    message("no metrics were recorded by any step")
  }
  jsonlite::write_json(out, "artifacts/metrics.json", auto_unbox = TRUE, null = "null")
})
print("pipeline finished")
```

## Metrics recorded by this run

| Metric | Value |
| --- | --- |
| n15::strains_recovered | 20 |
| n11::weighted_unifrac_permanova_R2 | 0.3693 |
| n7::samples | 20 |
| n14::tested | 314 |
| n11::unweighted_unifrac_spread_by_group | Early 0.195; Late 0.205 |
| n1::paired_end | True |
| n5::classified_phylum_pct | 100 |
| n14::source | MetaCyc pathways |
| n1::samples | 20 |
| n12::comparison | Late vs Early |
| n10::Simpson_median | 0.948 |
| n10::Shannon_median | 3.425 |
| n4::asvs | 229 |
| n11::weighted_unifrac_permanova_p | 0.001 |
| n5::classified_kingdom_pct | 100 |
| n10::Faith_PD_p | 0.05 |
| n12::significant | 5 |
| n1::median_reads_per_sample | 5913.5 |
| n1::min_reads_per_sample | 3178 |
| n4::samples | 20 |
| n0::rows | 20 |
| n13::asvs | 229 |
| n11::samples_without_group_left_out | Mock |
| n3::reads_kept | 139338 |
| n11::jaccard_dispersion_p | 0.319 |
| n13::asvs_excluded_by_nsti | 1 |
| n2::reads_in | 152360 |
| n10::Faith_PD_median | 9.339 |
| n10::Chao1_median | 83.5 |
| n9::top_taxon | Bacteroidota |
| n5::classified_family_pct | 90.8 |
| n3::reads_kept_pct | 91.5 |
| n10::Pielou_median | 0.804 |
| n15::asvs_shared_by_strains | 1 |
| n6::alignment_columns | 259 |
| n6::tips | 229 |
| n12::features_tested | 52 |
| n11::weighted_unifrac_spread_by_group | Early 0.104; Late 0.063 |
| n15::strains_missing | P.acnes |
| n13::ko_families | 5653 |
| n0::read_files | 40 |
| n11::unweighted_unifrac_dispersion_p | 0.707 |
| n11::jaccard_spread_by_group | Early 0.301; Late 0.330 |
| n5::classified_order_pct | 100 |
| n3::trunc_len_reverse | 159 |
| n5::reference | SILVA 138.1 (Zenodo 4587955), 16S |
| n14::comparison | Late vs Early |
| n10::Observed_median | 83.5 |
| n11::unweighted_unifrac_permanova_R2 | 0.4038 |
| n15::expected_strains | 21 |
| n12::method | ANCOM-BC2 |
| n11::unweighted_unifrac_permanova_p | 0.001 |
| n15::expected_sequences | 32 |
| n0::columns | 4 |
| n9::top_taxon_mean_pct | 63.5 |
| n13::weighted_mean_nsti | 0.213 |
| n7::total_reads | 122931 |
| n15::reads_not_exact_pct | 0 |
| n15::asvs_exact_match | 19 |
| n14::significant | 73 |
| n4::chimeric_reads_pct | 3.69 |
| n11::distances | Bray-Curtis, Jaccard, Unweighted UniFrac, Weighted UniFrac |
| n11::bray_curtis_dispersion_p | 0.005 |
| n5::classified_class_pct | 100 |
| n2::primers_absent | True |
| n3::truncation | from the quality profile (lower quartile above Q25) |
| n11::jaccard_permanova_R2 | 0.3154 |
| n7::features | 229 |
| n10::Simpson_p | 0.624 |
| n2::reads_out | 152360 |
| n3::trunc_len_forward | 243 |
| n1::fastqc_modules_failed | 176 |
| n4::reads_retained_pct | 80.7 |
| n14::method | ALDEx2 |
| n11::jaccard_permanova_p | 0.001 |
| n10::Pielou_p | 0.683 |
| n1::fastqc_modules_warned | 44 |
| n15::mock_sample | Mock |
| n13::metacyc_pathways | 341 |
| n4::reads_non_chimeric | 122931 |
| n10::Shannon_p | 0.121 |
| n2::primer_found_pct | 0 |
| n5::classified_species_pct | 6.1 |
| n10::Chao1_p | 0.165 |
| n9::rank | Phylum |
| n15::asvs_not_expected | 0 |
| n5::classified_genus_pct | 55 |
| n10::Observed_p | 0.165 |
| n11::bray_curtis_permanova_p | 0.001 |
| n11::weighted_unifrac_dispersion_p | 0.029 |
| n11::bray_curtis_permanova_R2 | 0.4324 |
| n3::reads_in | 152360 |
| n15::asvs_in_mock | 19 |
| n13::kegg_pathways | 145 |
| n13::ec_numbers | 1752 |
| n11::bray_curtis_spread_by_group | Early 0.224; Late 0.151 |
| n9::taxa_at_rank | 9 |

